# LURN… To Use R for Basic Calculus Tasks{#Calculus} 


```{r setup, include=FALSE, purl=FALSE} 
rm(list = ls())
if (!file.exists("Data")) dir.create("Data")
if (!file.exists("figures")) dir.create("figures")
if (!file.exists("tables")) dir.create("tables")
if (!file.exists("Other")) dir.create("Other")
set.seed(222664)
options(width = 60,  digits = 4, continue="   ")
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
Blind = file.exists("Blind.bat")
FigCap="To FIX"
``` 
 


```{r ChSetup, include=FALSE, purl=FALSE}
opts_chunk$set(fig.path='figures/') 
```  
 

This chapter explains how various tasks in introductory calculus can be
undertaken using R.

## Creating mathematical expressions

Before we embark on doing those fun things we want to do, we need to
learn how to store a mathematical expression as an object in R. There
are many keywords that appear in these expressions which are translated
to symbols when seen printed in a graph window but remain in ordinary
text in R commands. Such symbols include Greek letters, hats and bars,
all of which have a meaning.

## Plotting functions

This can be a lot of fun if you set things up right to start with. We
create a `function` to represent the mathematical expression
using the `function()` command, then plot it using either the
`plot.function()` or `curve()` commands. 


```{r SineCurve} 
MySine <- function(x){ sin(x)} 
curve(MySine, -4*pi, 2*pi, n = 2001) 
```  
 

[Link to SVG version](../SVGs/CalculusSineCurve-1.html)

You can use any valid mathematical expression in the
`function` you create. You might find it useful to test your
idea= values for your inputs that are easy to work with, such as zero
and one. You could then use integers, including negative integers, but
it could be easier to test your functions using values that lead to
identifiable results such as multiples of $\pi$ when working with
trigonometric functions. For example: 


```{r EvalMySine>> 
MySine(c((0:4)*pi/2)) 
```  
 

## Differentiation

The internal code used for differentiation in R is not as comprehensive
as other mathematical software. It can handle arithmetic operators `+`, `-`, `\*`, `/`, and `^`, and the single-variable functions `exp()`, `log()` (natural logarithm), `sin()`, `cos()`, `tan()`, `sinh()`, `cosh()`, `sqrt()`, `pnorm()`, `dnorm()`, `asin()` (inverse sine), `acos()` (inverse cosine), `atan()` (inverse tangent), and various `gamma()` functions. The functions `pnorm()` and `dnorm()` are restricted to the standard
normal distribution.

The easiest way to find the derivative of an expression is to store the
mathematical function using the `expression()` command, and then
use the `D()` or `deriv()` commands. 


```{r Derivation} 
y=expression(exp(2*x)) 
D(y,"x") 
```  
 

Nice that the answer comes out correct, albeit in a slightly different
way to what we might expect, because the answer has the multiplication
by 2 at the end instead of the more familiar $2e^{2x}$ wee expect to
see when explicit multiplication signs are not used. Also note that the
second argument to the `D()` function needs quote marks around
it. 

```{r cleanup, include=FALSE, purl=FALSE}
rm(list = ls())
``` 
 

