# LURN… To Create Complex Graphs{#GraphsWithExtras} 


```{r setup, include=FALSE, purl=FALSE} 
rm(list = ls())
if (!file.exists("Data")) dir.create("Data")
if (!file.exists("figures")) dir.create("figures")
if (!file.exists("tables")) dir.create("tables")
if (!file.exists("Other")) dir.create("Other")
set.seed(222664)
options(width = 60,  digits = 4, continue="   ")
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
Blind = file.exists("Blind.bat")
FigCap="To FIX"
``` 
 


```{r ChSetup, include=FALSE, purl=FALSE}
opts_chunk$set(fig.path='figures/') 
```  
 

This chapter assumes an understanding of the material in
Chapter \@ref(BasicGraphs) on creating simple graphs. That material
included adding simple elements of good graphing practice, like
meaningful titles and axis labels. In this chapter we see colour,
multiple graphs in one window, multiple sets of points on a single set
of axes, adding extra points or lines, and introduces colour, different
symbols and the legends that will then become necessary additions.

## More than one graph in a single window

We have already seen how to combine a set of graphs to be collected into
one window so that they form a single object for inserting into
documents. The `par()` command controls a large number of graphical parameters, including the layout of the graph window. One of
its common uses is to split up the graph window quickly.

Figure \@ref(fig:AirQuality-OzoneWindResidPlots) shows the residual analysis plots for a regression model. Four separate graphs form that residual analysis by default. If we do not combine them
into a single window using the `par()` command we need to
manipulate the four graphs separately when it comes time to include them
in a report.

The `mfrow` argument in the `par()` command tells R how to divide up the graph window. Note that it divides the existing
window so that the aspect ratio of the separate panels may change from
what you expect unless you either change the overall shape of the graph
window (see Figure \@ref(fig:AirQuality-BoxplotWind) where we did this for a
boxplot) or you divide the window up equally in both vertical and
horizontal directions (see Figure \@ref(fig:AirQuality-OzoneWindResidPlots) ). In
all examples presented thus far, the window has been divided up into
four equal sections. R then inserts successively created graphs into
those sections until the window has four graphs in it. A fifth graph
would appear on a new window that had four sections in it; this would
fit graphs five to eight, with the ninth going onto a new sheet and so
on. You can split the window into what ever size grid you like. Keeping
the aspect ratio of the graphs is a good idea, but making them slightly
wider than they are high is better than the reverse. That is, five rows
of four graphs is better than four rows of five graphs, unless you are
prepared to change the overall height and width of the window.

To illustrate creation of a lot of graphs and the efficiency of managing
them as a single graph window, we will create the histograms of twelve
distinct random samples of size 100, taken from the standard normal
distribution.

The final graph appears in Figure \@ref(fig:Hists12Z) , but there are a
couple of functions that need explaining. First, the `par()` command with the `mfrow` argument opens a graphing window and splits it into twelve sections. The second line uses the `for()` command to create a loop which is just a set of commands that are
grouped together and run over and over until the end of the loop. In
this case, the loop is run a total of twelve times with an arbitrary
label given to the counter used. A loop’s counter is often not used in
the processing of the code so is given a meaningless name; in this
instance, the word `Counter` is used to point out where the counter appears in the code. The third line creates a histogram with the
`hist()` command, of a set of 100 random values taken from the
standard normal distribution, found using the `rnorm()` command. Note the code also adds a title to each of the distinct graphs. This is
done using the `paste()` command which joins some text with the
counter number to form the title that will, as a consequence, change for
each graph. If there had been two or more commands inside the loop, they
would need to be grouped together using curly braces. Ignoring the
braces would mean only the first line of the set would be issued. The braces are included here as an example of good practice.

```{r include=FALSE}
FigCap="Twelve samples of size 100 drawn from the standard normal
distribution, plotted using separate histograms in a single window"
``` 

```{r Hists12Z, fig.cap=FigCap} 
par(mfrow=c(4,3)) 
for(Counter in 1:12){ 
hist(rnorm(100), main=paste("Sample", Counter))} 
``` 
 
 

﻿[Link to SVG version](../SVGs/ComplexGraphsHists12Z-1.html)

## Allowing different sized graphs to be included in a single window

The `layout()` command is very powerful, but can be a nightmare
to sort out. Before demonstrating how to implemet the command, I need to
show you what is possible by explaining how to think about the way the
command works.

Regardless of the size and shape of the particular graphs and how you
want them arranged, you need to think of the graphing window as a
rectangle that can be split into any number of rows and columns. The
shape of the sections of the window are dependent on the number of rows
and columns of course so for the time being let’s work with a
$4\times4$ grid. This means the 16 sections are the same shape as the
overall window.

We tell the `layout()` command how to use up the sections by
giving each section a label. The sections that get used for the first
graph will be given a “1", sections for the second graph get a “2" and
so on. For illustration, let’s use up nine of the 16 sections for a
large graph (a scatter plot), and place three smaller ones down the side
of it; below those, we will add a pair of boxplots which each need two
sections of the window.

The `layout()` command needs to interpret these labels as a
matrix. It would probably help to write out what you want for the
arrangement, create the matrix of labels, check that the two agree, and
then worry about the code for the various graphs. So for my description,
I want a $4\times4$ matrix, with a total of six different numbers in
it. Nine of them will be ones, a two, a three, a four, two fives, and
two sixes. They will be laid out as follows: $$\begin{array}
{cccc} 1&1&1&2\\ 1&1&1&3\\ 1&1&1&4\\ 5&5&6&6\\ \end{array}$$

Getting this matrix in R means identifying its contents, naming exactly
how many rows or columns are needed and how to insert the values into
the defined structure. We use the `matrix()` command to create the object for our arrangement, and the `c()` command to combine
the labels into a single object. Arguments to the `matrix()` command we need are either the `nrow` or `ncol` (one will do) and the `byrow` switch that tells the `matrix()` command to fill up the matrix by row until it is
full. 


```{r CreateArrangement} 
Arrangement.mat = matrix(c(1,1,1,2,1,1,1,3,1,1,1,4,5,5,6,6), byrow=TRUE, nrow=4) 
Arrangement.mat 
```  
 

﻿Having established the success of the code for the arrangement
structure, we just open the graph window using the `layout()` command, and add plots in the order our labels specify. The full set of
instructions is given in Figure \@ref(fig:Quakes-LayoutDiffSizes). Note that the
graphs are chosen for the illustration of the `layout()` command, not for the purposes of showing anything of interest in any of
the graphs created. The `quakes` data set distributed with the base installation of R is used in this illustration.

```{r include=FALSE}
FigCap="A selection of differing size graphs combined into one graph
window"
```

```{r Quakes-LayoutDiffSizes, fig.cap=FigCap} 
attach(quakes) 
layout(Arrangement.mat) 
plot(long, lat) 
plot(depth, mag, ylim=c(4,6.5)) 
plot(lat, mag, ylim=c(4,6.5), cex=0.5) 
plot(long, mag, ylim=c(4,6.5), cex=0.25) 
boxplot(mag, xlab="Magnitude", horizontal=TRUE) 
boxplot(log(mag), xlab="log(Magnitude)", horizontal=TRUE) 
detach(quakes) 
```  
 

﻿[Link to SVG version](../SVGs/ComplexGraphsLayoutDiffSizes-1.html)

Note that the way the points have been plotted and the way the text
appears in these graphs, differs. The addition of the `cex` argument in the third and fourth `plot()`  ommands forces the
size of the plotted symbol to be reduced. This is known as character
expansion , and can be applied to the text that appears on the graphs as well. Individual elements of text are adjusted using the `cex.main`, `cex.sub`, `cex.lab`  , and `cex.axis` arguments. See the help for the `par()` command for further details on their use.

## Using colour or different symbols

We show how to create a graph with different colours and symbols without
passing judgement on the suitability of the colours or symbols used. The
important feature to remember when you are about to create such a graph
is that the reader will not necessarily see the difference in colour as
you do. Similarly, the selection of symbol should be thought out
carefully. We do not want a solid symbol to appear bigger than a hollow
one or vice versa.

We use the `iris` data supplied with R for this example. It
has five variables in all; four are for numeric measurements and the
fifth is the species. In Figure \@ref(fig:IrisInColour) we use three
colours for the species and in Figure \@ref(fig:IrisInSymbols) we use different symbols. There is no reason why you might not use both arguments in combination to use different colours for the different symbols.

```{r include=FALSE}
FigCap="Sepal length and width plotted using different colours for
representing species. "
```

```{r IrisInColour, fig.cap=FigCap} 
plot(Sepal.Length~Sepal.Width, data=iris, xlab="Sepal width", ylab="Sepal length", col=Species) 
```  
 

[Link to SVG version](../SVGs/ComplexGraphsIrisInColour-1.html)

```{r include=FALSE}
FigCap="Petal length and width plotted using different symbols for
representing species. "
```

```{r IrisInSymbols, fig.cap=FigCap} 
plot(Petal.Length~Petal.Width, data=iris, xlab="Petal width", ylab="Petal length", pch=as.numeric(Species)) 
```  
 

[Link to SVG version](../SVGs/ComplexGraphsIrisInSymbols-1.html)

Note that we needed to convert the species names to numbers using the
`as.numeric()` command to get the symbols to be plotted.

R seems to have converted the species names to numbers when we called
for different colours, starting with 1 for black, 2 for red, and 3 for
green. As an aside, the colour can be specified by name or by a number.
There are well over 600 named colours to choose from so the full list is
therefore not printed here. Type 


```{r Colors, eval=FALSE} 
colors() 
```  
 

﻿if you are really keen to know what they all are. If this list is still
not sufficient to meet your needs, you can design your own colour using
a combination of red, green, and blue components. This is explained
under the help for the `par()` command.

## Adding points to an existing graph

## Using lines instead of points {#LinesInsteadOfPoints}

The default action for any scatter plot is to mark each point. If the
set of *x*-values are ordered we can create a line plot. This is often
seen as a time series plot. We create a time series plot very easily
using the `plot()` command on an object that has been stored as
a time series. There are many examples of time series in the
`datasets` package. We use the `ldeaths` series which counts the number of deaths due to lung diseases in the United
Kingdom, and its associated series for male and female deaths.

In Figure \@ref(fig:LungsNoLegend) , we see a set of three commands. The
first plots a time series on a set of axes and the axis labels. The
second and third each add a set of line segments from the other two
series using the `lines()` command.

```{r include=FALSE}
FigCap="Deaths from lung diseases in the United Kingdom; there is no legend (on purpose)."
```

```{r LungsNoLegend, fig.cap=FigCap} 
plot(ldeaths, ylim=c(0,4000), xlab="Year", ylab="Number of deaths") 
lines(fdeaths, col="red") 
lines(mdeaths, col="blue") 
```  
 

﻿[Link to SVG version](../SVGs/ComplexGraphsDeathsTSPlot-1.html)

We will see that the plot can be improved by adding a legend in
Figure \@ref(fig:LungsLegend).

## Adding lines to an existing graph

In Section \@ref(AddFittedLine) we saw how to add a fitted line from a
simple regression model to a plot of the data used to fit the model,
using the `abline()` command. This command has other uses that
are explained here. This section is only interested in the addition of a
single straight line at a time; a series of line segments joined
together was covered in Section \@ref(LinesInsteadOfPoints).

Perhaps the easiest tasks of adding lines to a plot is to add either a
horizontal or a vertical line. In Figure \@ref(fig:QuakesPlaces) , we see
the locations of the quakes marked by their latitude and longitude. On
this occasion, the graph window has been left square and the spacing on
the graph itself would be different for the latitude and longitude if we
did not force the axes to have similar length for each direction. The
`xlim` and `ylim` arguments create the axes, but
note the need to correctly define the axis for latitude from a large
negative number to the less negative number so that the orientation of
north and south is kept consistent with convention. Fiji is in the
southern hemisphere after all\!

The `plot()` command creates the axes and plots the points for
the earthquakes in Figure \@ref(fig:QuakesPlaces). We then add two
vertical and two horizontal lines to the plot, both using the `abline()` command.

```{r include=FALSE}
FigCap="Horizontal and vertical lines added to a scatter plot"
```

```{r QuakesPlaces, fig.cap=FigCap} 
plot(lat~long, data=quakes, xlim=c(160,190), ylim=c(-40,-10)) 
abline(h=-30) 
abline(h=-20, lty=1) 
abline(v=180, lty=2) 
abline(v=170, lty=3) 
```  
 

﻿[Link to SVG version](../SVGs/ComplexGraphsQuakesPlaces-1.html)

Notice that the four commands used to add the four extra lines state the
horizontal or vertical placement with a single letter, and that the line
type has been chosen in three instances. It turns out that line type 1
(a solid line) is the default. More types do exist but those given are
the most commonly chosen. The thickness of the lines can be altered
also, using the `lwd` argument. The default line width is 1.

It is also possible to add a straight line that is niether vertical or
horizontal using the `abline()` command. To achieve this we need
to know the slope and intercept of the line. This is the form of this
command used when we added the fitted line for a simple linear
regression model to the plot in Section \@ref(AddFittedLine). The
plotted line will have the equation $y=a+bx$ and we need only supply
the values of `a` and `b` to the `abline()` command. 

## Adding a curve to a graph {#AddCurves}

Use of `abline()` is great if you have a straight line to add to
a plot, but as soon as the line needs to be curved the problems begin.
We will see in Figure \@ref(fig:HistWithNormalCurve) that it isn’t actually too difficult if we use the
`curve()`  command. First, we add a curve representing the normal
probability density function to a histogram of random values. 

```{r include=FALSE}
FigCap = "A normal density curve added to a histogram."
```


```{r HistWithNormalCurve, fig.cap=FigCap} 
hist(rnorm(1000)) 
curve(dnorm(x), -3,3, add=TRUE) 
```  
 

﻿[Link to SVG version](../SVGs/ComplexGraphsHistWithNormalCurve-1.html)

The `curve()` function has four arguments in this example. First
is the function being plotted, which in this case is `dnorm()` —the density function for the normal distribution. We then supply the
range of values over which this function is to be evaluated using two
arguments, and finally we ask that the curve be aded to the existing
plot. There are other arguments that we could add, including
`n` , the number of points to be evaluated over the specified
range.

The second example shown in Figure \@ref(fig:AddPolyCurve) below  is the addition of the quadratic fitted
line for a regression model. We use the example given in Section \@ref(PolyReg). 


```{r include=FALSE	}
FigCap = "The quadratic curve for the fitted values from a polynomial regression model added to the scatter plot of the data."
```

```{r AddPolyCurve, fig.cap=FigCap} 
Ozone.poly2 = lm(Ozone~poly(Wind,2, raw=TRUE), data=airquality) 
Coeffs=coef(Ozone.poly2) 
plot(Ozone~Wind, data=airquality) 
curve(Coeffs[1]+Coeffs[2]*x+Coeffs[2]*x^2, min(airquality$Wind), max(airquality$Wind), add=TRUE) 
```  
 

[Link to SVG version](../SVGs/ComplexGraphsAddPolyCurve-1.html)

Note that the `poly()` function needs to use the `raw=TRUE` argument to get the correct coefficients for the curve to be added to this graph.

## Adding text to an existing graph

## Adding a legend for different information within a graph

```{r include=FALSE}
FigCap="Deaths from lung diseases in the United Kingdom (with legend
added)"
``` 
```{r LungsLegend, fig.cap=FigCap} 
plot(ldeaths, ylim=c(0,4000), xlab="Year", ylab="Number of deaths") 
lines(fdeaths, col="red") 
lines(mdeaths, col="blue") 
legend("topright", legend=c("Total", "Men", "Women"), col=c("black", "blue", "red"), lty=c(1,1,1)) 
```  
 

﻿[Link to SVG version](../SVGs/ComplexGraphsDeathsTSPlotLegend-1.html)

## More complex bar charts

A bar chart, as seen in Section \@ref(BarCharts) , has a bar showing
the frequency of each level of a single factor. A cluster bar chart has
a bar for each combination of two factors. The clustering indicates how
the bars are grouped together. A stack bar chart on the other hand has a
single bar for each level of one factor split into sections for each
level of the second factor; it therefore appears to stack the bars on
top of one another.

To create an example, we will first create a list of frequencies of
outcomes and the two categorical variables required. The cluster bar
chart and stack bar chart then appear in
Figures \@ref(fig:ClusterBarChart) and \@ref(fig:StackBarChart)
respectively. Our data consist of the counts of males and females of
ages 17 to 20. 


```{r CreateSomeData} 
Frequency=c(12,19, 22, 18, 11,13, 13, 12) 
Gender=rep(c("Male", "Female"), 4) 
Age=rep(c(17,18,19,20), each=2) 
```  
 

To create the two types of bar chart, we need to create a matrix for the
counts to be used in the graphs. In this instance, this is done using
the `tapply()` command. The actual plotting is done using the
`barplot()` command which recognizes it is working on this
matrix and acts accordingly, including the ability to generate a legend
for the graphs.

```{r include=FALSE}
FigCap="An example of a cluster bar chart"
```

```{r ClusterBarChart, fig.cap=FigCap} 
barplot(tapply(Frequency, list(Gender, Age), sum), col=c(2,4), legend=TRUE, beside=TRUE) 
title("Cluster bar chart of age and gender", ylab="Count", xlab="Age") 
```   
 

[Link to SVG version](../SVGs/ComplexGraphsClusterBar-1.html)

```{r include=FALSE}
FigCap="An example of a stack bar chart"
``` 

```{r StackBarChart, fig.cap=FigCap} 
barplot(tapply(Frequency, list(Gender, Age), sum), col=c(2,4), legend=TRUE) 
title("Stack bar chart of age and gender", ylab="Count", xlab="Age") 
```   
 

﻿[Link to SVG version](../SVGs/ComplexGraphsStackBar-1.html)

## Contour plots

A scatter plot indicates the presence of an observation at pairs of
*x* and *y* values. When a third variable is also measured, we run
into some trouble when comparing it to the values of *x* and *y* in
a single graph. Three-dimensional displays can be employed but require
the interpretation of the creator and the reader to turn the
two-dimensional graph on paper or the screen into the three-dimensional
object actually being displayed.

A contour plot is a two-dimensional display. It shows a set of curves
that represent common values of a variable plotted against two other
variables. They are commonly seen in weather charts where barometric
pressure lines (contours) are plotted on the map of a country or region
(latitude and longitude are the two other variables).

Figure \@ref(fig:MtEdenContour) shows the contour lines for the altitude
of a volcano in Auckland, New Zealand. Mt Eden is also known by its
Maori name Maunga Whau and the data we need for this illustration is
provided in the datasets supplied with base installations of R.

```{r include=FALSE}
FigCap="Contour plot of Mt Eden, an extinct  volcano in Auckland, New
Zealand."
``` 

```{r MtEdenContour, fig.cap=FigCap} 
contour(volcano) 
```  
 

﻿[Link to SVG version](../SVGs/ComplexGraphsMtEdenContour-1.html)

This provides a black and white image that has the contours marked with
the value of the third variable. A coloured version of the plot is
available using the `filled.contour()` command in place of the
`contour()` command in Figure \@ref(fig:MtEdenContour). An even
more impressive version of this plot can be obtained by looking at the
example that accompanies the help for the `volcano` data.
Typing: 


```{r VolcanoExample, eval=FALSE} 
example(volcano) 
```  
 

﻿generates the plot using the code given in the help file.

The `volcano` data is stored as a matrix of altitudes. The
`contour()` command uses this to interpolate the coordinates of
the points where the altitude is constant.

A further example is generating the contours for a function. Let’s plot
the function $$z=0.75x^2 + 0.5y^2 -0.4xy + x - y -1$$ over the range
$-5\leq x,y \leq 5$. All necessary codes are given in Figure \@ref(fig:XYZContour) which includes the contour plot of the
function.

```{r include=FALSE}
FigCap="Contour plot showing values of *z*, where
$z=0.75x^2 + 0.5y^2 -0.4xy + x - y -1$"
```

```{r XYZContour, fig.cap=FigCap} 
xUnique=((0:100)-50)/10 
yUnique=((0:100)-50)/10 
x=rep(xUnique, 101) 
y=rep(yUnique, each=101) 
z=matrix(0.75*x^2 + 0.5*y^2-0.4*x*y+x-y-1, nrow=101) 
contour(xUnique, yUnique, z) 
```  
 

[Link to SVG version](../SVGs/ComplexGraphsXYZContour-1.html)

Note the `contour()`  function needed to know the unique values
of the *x* and *y* variables we created, but that the *z* matrix
needed these values repeated enough times so that the function was
evaluated over the entire set of combinations of *x* and *y*. 

```{r cleanup, include=FALSE, purl=FALSE}
rm(list = ls())
``` 
 

