# LURN… To Export a Dataset{#ExportingData} 


```{r setup, include=FALSE, purl=FALSE} 
rm(list = ls())
if (!file.exists("Data")) dir.create("Data")
if (!file.exists("figures")) dir.create("figures")
if (!file.exists("tables")) dir.create("tables")
if (!file.exists("Other")) dir.create("Other")
set.seed(222664)
options(width = 60,  digits = 4, continue="   ")
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
Blind = file.exists("Blind.bat")
FigCap="To FIX"
``` 
 

﻿This chapter assumes you’ve got some data to export, and I mean data;
not results from analyses which is discussed in
Chapter \@ref(StoringResults). You may also wish to use this
information once you know how to tabulate various numerical summaries,
discussed in Chapter \@ref(NumericalEDA).

## Creating external files

This section is just the inverse action of importing data from an
external file as discussed in Chapter \@ref(ImportingData). All the
`read` type commands have corresponding `write` commands.

If you read Chapter \@ref(ImportingData) you’ll know that my preference
is to use files that are easily transferred and easily checked for their
accuracy using other software. In particular, I prefer to create comma
separated value (csv) files using the `write.csv()` command. If
there was a `data.frame` called `Chickens` in my
current workspace that I wanted to export for sharing with another user,
I would issue the following command. 


```{r WriteCSV, eval=FALSE} 
write.csv(Chickens, file="chickens.csv") 
```  
 

If R had created rownames for the `data.frame` , I might not
want to include them. A simple additional argument is all that is
required. 


```{r WriteCSV2, eval=FALSE} 
write.csv(Chickens, file="chickens.csv", row.names=FALSE) 
```  
 

If another file type is required then the help page for
`write.csv()` may be useful. This page also gives the help on
the `write.table()` command.

## Exporting data for use in alternative software

The better statistical software options allow users to import data from
a range of sources and file types. If for some (very strange) reason you
need to create files in a specific format for importing into another
statistics package then you will need to investigate the
`foreign` package. 

```{r cleanup, include=FALSE, purl=FALSE}
rm(list = ls())
``` 
 

