
# LURN… To Import Data {#ImportingData} 


```{r setup, include=FALSE, purl=FALSE} 
rm(list = ls())
if (!file.exists("Data")) dir.create("Data")
if (!file.exists("figures")) dir.create("figures")
if (!file.exists("tables")) dir.create("tables")
if (!file.exists("Other")) dir.create("Other")
set.seed(222664)
options(width = 60,  digits = 4, continue="   ")
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
Blind = file.exists("Blind.bat")
FigCap="To FIX"
``` 
 

﻿This chapter covers the methods required to pull data from external
sources into R. If you want to create data within R , you should be
reading Chapter \@ref(DataEntry).

## Data from external files

While I prefer to use files extracted from EXCEL with comma delimited
values, R handles many common formats such as plain text files with
space or tab delimiters. You need to know what format a file is,
probably by opening it and actually seeing if it is as you expect. It is
too easy to rename a file with various extensions which may have
meanings in your operating system that have little relevance to R. A
specific case in point is when you are presented with a file having the
extension, which is commonly assumed to be a text file. We need to know
if the first line of information in the file is actual data or the
headings for the columns of data. We also need to know what symbol is
used to separate the columns of the data; spaces, commas, or tabs are
the most common options. Each of these options has a distinct R command
associated with it, but all of these commands link back to the same
`read.table()` command.

The various commands are as follows

| Delimiter | R command              | Common Extensions |
|:-------- |:--------------------- |:---------------- |
| space     | `read.table()` | DAT, TXT          |
| tab       | `read.delim()` | TXT               |
| comma     | `read.csv()`   | CSV, TXT          |

Note that the extension `txt` appears as possible extensions for all delimiter
types. R will not assume any extension for these commands. You will need
to explicitly state the full filename including extension when using
these commands. Some extensions will have a default program associated
with them by your operating system. For example, files will be opened in
under Windows, and if Microsoft Office is installed on your machine, a
file will be opened in Microsoft Excel.

To import a comma delimited file called `chickens.csv`, you would issue the following command 


```{r ImportCSV, eval=FALSE} 
Chickens = read.csv("chickens.csv") 
```  
 

In this most basic form, the `read.csv()` command will look for
the file in the current working directory, and import it into R and
store it as a `data.frame` called `Chickens`. The
default settings of the `read.csv()` command are to have a
`header` row in the file and to have no `row.names` attribute associated with the data in the file. If your data file
already had a column for the names of the chickens as the first column,
you would issue the command 


```{r ImportCSV2,eval=FALSE} 
Chickens = read.csv("chickens.csv", row.names=1) 
```  
 

and if the data did not currently have any column headings you would
issue the command 


```{r ImportCSV3,eval=FALSE} 
Chickens = read.csv("chickens.csv", header=FALSE) 
```  
 

There are other settings to consider which you can investigate using the
help for the `read.csv()` command by typing
`?read.csv` at the command prompt. This help page is actually
a combined help page for the family of commands described in this
section.

The working above assumed you could put the data file into the correct
working directory. So where was that? To find out where R thinks you are
currently working, use the `getwd()` command. Note that the
output may look a little strange to some users, especially Windows
users. What do I mean? Look at the following: 

```{r GetWD} 
getwd() 
```
  
 

The full path to the working directory where this chapter was processed
has been displayed, starting with the letter associated with the hard
drive, followed by a colon. Then the fun begins. The folder structure is
represented using forward slash signs, not the backslash used in Windows
operating systems even though the processing of this work is done using
a Windows machine. The rest should be as expected and you could find the
right folder by looking in the appropriate place on your hard drive. The
reason for R ’s use of the forward slash is not entirely simple to
explain, but in short it is because the standard backslash symbol has a
special use in R. For the moment, the choice of slash versus backslash
is not important. It is important when we need to type out the path to
the location of a file for ourselves.

## Data stored in another directory

I do try to keep each separate project in its own distinct directory,
and moving the raw data file to that directory makes sense. It does not
make sense to have multiple copies of a dataset though, so we need to
know how to pull a file from a different working directory into R.

When I displayed the current working directory in the previous section
using the `getwd()` command, we saw the way that R used forward
slash symbols to denote the hierarchy of folders on our hard drive.

If we know the complete specification of the location of a file, right
the way from the name of the hard drive down the directory tree to the
actual location, the way we specify the location needs to match the way
R has printed a path. That is, we use the forward slash symbol not a
single backslash. If we do want to use a backslash symbol, we would need
to use a double backslash, not just a single one. It is better to use a
single forward slash symbol however. This is because the single forward
slash presentation works for all operating systems and means our code
can be shared to users of all operating systems. You might not plan to
do this right now, but let’s use good habits from the start.

It is common for data to be stored in a folder that is close to the one
we are working in. We might have a folder called which is within the
working directory (a subfolder), or it might be a folder at the same
level in the directory tree as the current working directory. In either
case, we don’t need to specify the location using the full path. The
term used to describe the location is a relative path because the
reference to the location is relative to the current location having
focus (our working directory).

If our data set is stored in the file in a subfolder called , we can use 


```{r ImportCSVSubFolder1, eval=FALSE} 
Chickens = read.csv("MyData/chickens.csv") 
```  
 

to pull it into our current workspace. This is actually shorthand for
the more complete form 


```{r ImportCSVSubFolder2, eval=FALSE} 
Chickens = read.csv("./MyData/chickens.csv") 
```  
 

where the current folder is denoted using the single period followed by
the first slash. Personally, I would prefer to see the more complete
form of this relative path being used but it is personal preference
only.

If the folder was not a subfolder, but was on the same level in the
directory tree as our current working directory, we would use the
shorthand symbol for the parent directory (one level up the directory
tree). This is done using a double period. 


```{r ImportCSVSOtherFolder, eval=FALSE} 
Chickens = read.csv("../MyData/chickens.csv") 
```  
 

﻿Relative paths are therefore quite useful because they avoid having to
type out long paths. The full specification of apath might also become a
problem when files for a project are moved from one storage device to
another. For example, you might want to take your work to a friend,
tutor, or colleague on a memory stick or other portable storage medium.
Relative path referencing makes your work very transferable and
transportable.

## Data saved by other statistical software

Many statistical software packages use their own file types for storing
data. R is no different actually\! The chief problem we have is to find
a way of transferring data from one application to another. Like most
other statistics programs, R doesn’t handle all other file types. Some
files can be imported into R using the commands in the
`foreign` package, but it is probably best just to avoid the
problems from the start.

In many instances it will prove easiest to use copy and paste
functionality within your operating system to take the data from
whatever original source it was given to you in and put it into a
suitable spreadsheet program. Then save it using the comma separated
values format and read the data into R using the commands given in the
previous section.

I recommend trying to obtain the data in an easily imported file type
rather than attempting to use the functions in the `foreign` package.

## Data from files stored on the internet

Sometimes a data set is made available via the internet. If you can
obtain the full URL for the downloadable data file then it can be
entered into the `read.csv()` or `read.table()` commands. This exercise is seldom necessary except for data files that
you know will be updated for distribution through the web. Some
government agencies and financial database services do this.

## Data contained in contributed packages {#DataFromPackages}

The base installation of R includes a package called
`datasets`. These data sets are useful for testing code and
writing examples for insertion into documents like this one. Data sets
contained in the `datasets` package are actually ready and
waiting to be accessed, but often we want to bring the data into our
current workspace using a command such as 


```{r PullData} 
data(airquality) 
str(airquality) 
ls() 
```  
 

The `data()` command looks in the `datasets` package [@Rpkg-base] by default. If we wanted to get some data from another package we would
need to state the name of the package explicitly. For example, 


```{r PullData2} 
data(anorexia, package="MASS") 
```  
 

﻿(The `MASS` package is already installed by default.)

Often we will be using a particular data set because it is good for
demonstrating functions within a particular package. If we are loading
the package using the `library()` command, to get access to the
functions, we will have also made the data available. This is why the
data from the `datasets` package is ready for use; this package is
loaded by default whenever R is started.

## Data cleanliness

Rather ironically, the suggestion for including this section came from
my mother. All too frequently, data are entered by people who will not
actually need to use the data themselves. Sometimes data are entered by
different people and then compiled into a single dataset and the various
codes that some sources of data may use are not necessarily in common
with all other users. Take for example, the many ways we might enter
data for the gender of individuals answering a survey. You might easily
realise that an “`M` " means male, but R needs consistency.

It is extremely important that we have knowledge of the format of the
data we import. Use of the `str()` command is a good start, and
use of the `head()` command might also be useful.

Here are some pointers to look out for:  - Have you mixed the case of the text in your data? Remember that R is    case sensitive and that “R" is not the same as “r" in R.  - Has R tried to be too smart and interpreted plain text information    as if you wanted a `factor` instead? This occurs    frequently with date values which are notorious for the variety of    formats people choose. If you have a factor when you wanted just    plain text, investigate the `as.character()` command which    changes the format to being character valued data.  - Have a multitude of codes been used for variables like gender? If    so, R will decide that the variable is a `factor` and list    the different values the variable can take. The levels of the factor    can be re-assigned using the `levels()` command. See how    this is done on and note that two elements of the assignment can be    the same if the current codes mean the same thing.  - Have extra spaces crept into the codes for levels of factors    somehow? This happens when an extra space occurs between words or is    added on the end of a text string. Often, this problem is difficult    to spot visually as we are talking about white space — R will find    it though\! Use the `levels()` command to fix this if the    problem is small, but consider using a find/replace tool in a    spreadsheet application as an alternative.  - Have numbers come into R as text for some reason? This happens for a    number of reasons. A blank space being entered in the cell instead    of a code for a missing value for example. R would see this as a    character string “ ", not an empty cell. Another reason is that the    missing value code is textual and not recognized by R. There are    ways around this problem using either find/replace techniques in    your spreadsheet application to delete the missing value codes, or    by changing the way the `read.table()` or    `read.csv()` command reads your data file. See the help for    these commands if this is the case.

## Other packages

It is all too common to need to import data from a spreadsheet
application where the data is not conveniently placed in the upper-left
corner of the first sheet. There are a number of such spreadsheet
applications, but the most commonly used one is Microsoft Excel.

Data in spreadsheets is seldom ready for importing in the exact form we
want it. If you need to extract a set of data that is in a named sheet
or always appears in the same numbered sheet of a Microsoft Excel file,
then you might investigate the `openxlsx` package. See
Chapter \@ref(Additional) and the help page for the
`read.xlsx()` command. 

```{r cleanup, include=FALSE, purl=FALSE}
rm(list = ls())
``` 
 

