# LURN… To Use R for Linear Algebra{#LinearAlgebra} 


```{r setup, include=FALSE, purl=FALSE} 
rm(list = ls())
if (!file.exists("Data")) dir.create("Data")
if (!file.exists("figures")) dir.create("figures")
if (!file.exists("tables")) dir.create("tables")
if (!file.exists("Other")) dir.create("Other")
set.seed(222664)
options(width = 60,  digits = 4, continue="   ")
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
Blind = file.exists("Blind.bat")
FigCap="To FIX"
``` 
 


```{r ChSetup, include = FALSE, purl=FALSE}
opts_chunk$set(fig.path='figures/Algebra', out.width="0.7\\textwidth") 
```  
 

This chapter explains how some tasks commonly taught in introductory
algebra courses can be undertaken using R. The functionality is built
into R because so many tasks in statistical analyses require
manipulation of vectors and matrices, although much of this work is
hidden behind other commands and functions.

## Basic notes on storage of vectors and matrices

We will concern ourselves with only numeric valued vectors and matrices
in this chapter. Other types of data can be stored in vectors and
matrices (as well as other data structures) but they have little
relevance for linear algebra at this time.

In linear algebra, a vector is a one dimensional set of numbers. It is
usually referred to as a row vector or a column vector to indicate its
orientation. This orientation does matter for the ability to perform
certain tasks when working with two or more vectors and matrices.

A matrix is a two-dimensional array of numbers, having a number of rows
and columns. Many mathematical software applications do not explicitly
distinguish a vector from a matrix because they store a row vector as if
it were a matrix having just one row. Similarly, a column vector is
stored as a matrix having just one column.

R does distinguish betweena vector and a matrix, but it does not state
whether a vector is a row or column vector. This does have its
advantages, and probably some disadvantages as well. If for any reason
you need to force the orientation of a vector, then create it as a
matrix with one of its dimensions set equal to one. For example, we
create a simple vector: 


```{r CreateVector} 
x=1:4 
x 
class(x) 
```  
 

We can convert this to a column matrix using 


```{r CreateMatrix} 
X=matrix(x, ncol=1) 
class(X) 
X 
```  
 

Notice that I’ve used a capital letter for the matrix version of the
same set of numbers here which is possible because R is case sensitive.
The `ncol` argument can be replaced by using the
`nrow` argument if a row vector is required.

## Creating some simple matrix structures

The identity matrix is a diagonal matrix so we use the `diag()` command to create it. For example 


```{r CreateIdentity} 
I3=diag(3) 
I3 
```  
 

We often need to find the transpose of a matrix. The `t()` command does this. 


```{r Transpose} 
M=matrix(1:6,nrow=2) 
M 
t(M) 
```  
 

A Hilbert matrix can be created using a formula. The `row()` and
`col()` commands use the size of the argument to create objects
of the same size. 


```{r CreateHilbert} 
row(I3) 
Hilb3=row(I3)+col(I3)-1 
```  
 

Notice that the Hilbert matrix is created by adding two matrices
together and then a constant is subtracted from all elements of the
result.

## Matrix and vector calculations

We saw a simple addition of elements of two matrices of the same size as
well as a subtraction in the previous section. R uses the standard
arithmetic operators on pairs of matrices (or vectors) on an
element-by-element basis.

Addition of a constant to a vector or matrix is effected on all elements 


```{r AddConstant} 
x+2 
2+x 
```  
 

Similarly, we can multiply a vector or matrix by a scalar using: 


```{r ScalarMult} 
5*x 
x*5 
5*X 
```  
 

which retains the size of the vector or matrix. Also note that
pre-multiplying by a scalar is equivalent to post-multiplying by a
scalar, as should have been expected. This is of course different to
matrix multiplication.

Matrix operations use different symbols from simple arithmetic operators
to distinguish the different results that might be desired. For example, 


```{r MatrixElementMult} 
A=matrix(1:4, nrow=2) 
B=matrix(7:10, nrow=2) 
A*B 
```  
 

creates a matrix with elements based on simple multiplication of the
paired elements, not those based on proper matrix multiplication.

If we want to multiply two matrices, the order is important. 


```{r MatrixMult} 
A%*%B 
B%*%A 
```  
 

Multiplying a matrix by a vector is easily achieved, and R will orient
the vector to conform to the rules of matrix multiplication as required. 


```{r MultiplyMatrices} 
y=c(3,4) 
A%*%y 
y%*%A 
```  
 

## Inverting a matrix

The inverse of a matrix is used in many statistical procedures. The
`solve()` command inverts a square matrix. 


```{r GetInverse} 
A 
solve(A) 
```  
 

For matrices that are not square, a generalised inverse can be found
using the `ginv()` command found in the `MASS` package
that is part of the standard R distribution. 


```{r GetGInverse} 
M 
require(MASS) 
ginv(M) 
```  
 

Of course, the product of the matrix and its generalised inverse should
be an identity matrix. 


```{r ShowIdentity} 
M%*%ginv(M) 
```  
 

Like many other applications, R does not recognize the number zero very
well. This is due to the way decimal numbers are stored.

## Solving a set of linear equations

If we want to solve the set of equations, represented in matrix form as
$Ax=b$, we use the `solve()` command 


```{r SolveLinearSet} 
A 
b=c(5,7) 
solve(A,b) 
```  
 

## Determinants and traces

The `det()` command is used for finding the determinant of a
matrix. 


```{r Determinant} 
A 
det(A) 
```  
 

The trace of a matrix is not so easily found. There is a command that is
called `trace()` but this is not an algebraic function. If you
want the trace of the matrix, you will need to sum the eigenvalues of
the matrix.

## Eigenvalues and eigenvectors

Calculation of eigenvalues and their associated eigenvalues is fairly
simple, but extracting the elements required might pose a little
difficulty. Let’s create a 4$\times$4 Hilbert matrix and find its
eigenvalues: 


```{r FullEigen} 
H4=matrix(-1,nrow=4, ncol=4) 
H4=H4+row(H4)+col(H4) 
H4 
eigen(H4) 
```  
 

If we only wanted the eigenvalues, we could add the argument
`only.values=TRUE` ; if the eigenvectors were of interest,
then we would need to extract the second element of the above result: 


```{r EigenVectors} 
eigen(H4)$vectors 
```  
 

```{r cleanup, include=FALSE, purl=FALSE}
rm(list = ls())
``` 
 

