# LURN… To Manipulate your Datasets{#ManipulatingData} 


```{r setup, include=FALSE, purl=FALSE} 
rm(list = ls())
if (!file.exists("Data")) dir.create("Data")
if (!file.exists("figures")) dir.create("figures")
if (!file.exists("tables")) dir.create("tables")
if (!file.exists("Other")) dir.create("Other")
set.seed(222664)
options(width = 60,  digits = 4, continue="   ")
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
Blind = file.exists("Blind.bat")
FigCap="To FIX"
``` 
 


```{r ChSetup, include=FALSE, purl=FALSE}
opts_chunk$set(fig.path='figures/Manip', out.width="0.7\\textwidth") 
```  
 

﻿This chapter explains how to rearrange your data to meet your needs.
Basic sorting, data extraction and re-combination are discussed.

To get the most of this chapter, you should have worked through
Chapter \@ref(DataEntry) on entering data manually. You do not need
to have completely read through Chapter \@ref(ImportingData) on importing
data as the examples here do not rely on its content. Ultimately, you
will want to use the material in Chapter \@ref(ImportingData) in
conjunction with what follows here.

This chapter starts using the functionality that is built into R , but
we also take a look at an add-on package called `dplyr` [@Rpkg-dplyr]that is becomeing more and more commonly used.

## Sorting the data into a different order {#DataSorting}

The `sort()` command is illustrated in this section. It works
for sorting a single vector of numeric values into ascending or
descending order. 


```{r Sorting} 
x=round(rnorm(10),2) 
sort(x) 
sort(x, decreasing=TRUE) 
```  
 

We use the `order()` command to work with data that are
non-numeric vectors, as well as data in `matrix` or
`data.frame` format. For example, we first sort the names of
me and my three children initially given in descending order of age. 


```{r Ordering} 
Home=c("Jonathan", "Callum", "Annabelle", "Cordelia", "Hershey") 
order(Home) 
Home[order(Home)] 
```  
 

﻿So we see that the `order()` command doesn’t actually do the
re-ordering after all, but that we need to use the subscripts to extract
the elements (the full set) from the pre-existing set.

We do need to take note of how the upper and lower case letters will be
ordered. 


```{r Ordering2} 
SomeLetters=c("A", "B", "a", "c", "b", "C") 
SomeLetters[order(SomeLetters)] 
```  
 

The `order()` command can prove a useful method of re-ordering
experimental data if the run order has been stored as a variable. To
illustrate this point we create a `data.frame` with two
elements. 


```{r Ordering3} 
x=round(rnorm(10),2) 
x 
RunOrder=sample(10) 
```  
 

Note that the `sample()` command used here has created a random
permutation of the numbers one to ten. 


```{r Ordering4} 
RunOrder 
Data=data.frame(x,RunOrder) 
Data[order(RunOrder),] 
```  
 

## Extracting data using information on one variable

To illustrate the extraction of a set of observations from a
`data.frame`, we make use of one of the internal datasets
provided with the base installation of R. We retrieve the `airquality` data using: 


```{r GetData} 
data(airquality) 
```  
 

﻿An explanation of this command is given in
Section \@ref(DataFromPackages). This data set has its own help page
that can be viewed by typing `?airquality` at the R prompt.

We can ask for the rows of this data that match specified details using the `subset()` command. Various operators exist for numerical comparisons; the
simplest of which are `<` and `>`. For the purposes of brevity in presenting the results of the following examples
that follow, we use the `nrow()`  command to show how many
observations in the data meet the specified conditions. The following
commands extract the number of days where the wind speed was less than
ten miles per hour and then the temperature was less than 60 degrees Fahrenheit. 


```{r GrabRows1} 
nrow(subset(airquality, Wind<10) )
nrow(subset(airquality, Temp<60)) 
```  
 

If we want to know how many observations have an exact value, we must use
`==`,  not a single equals sign because it has a
different meaning in R. The double equals sign is the one used for
comparisons. For example, the number of days when the temperature was
recorded as 69 degrees can be found using: 


```{r GrabRows3} 
nrow(subset(airquality, Temp==69)) 
```  
 

## Extracting data using information on more than one variable

To print out the rows of the `airquality` `data.frame` where the temperature was less than 60 degrees
and the wind was less than ten miles per hour, we would use the
`&`  operator between the two conditions that
identify which rows are required. 


```{r GrabRows2} 
subset(airquality, Wind<10 & Temp<60)
```  
 


## Use of the dplyr package for data manipulation {#DPlyRDataManipulation}

The `dplyr` package [@Rpkg-dplyr] is an alternative to the basic R functionality shown in this chapter. The package creator (Hadley
Wickham) wants his package to make data manipulation easier for the
end-user, and more efficient with respect to computation time. He has
simplified data manipulations to a small set of commands:
`filter()`, `arrange()`, `sample\_n()`, `sample\_frac()`, `mutate()`, and `transmute()` are discussed here, while the others will be 
discussed in Section \@ref(DPlyRDataSummarisation) where they are more
relevant.

You’ll need to install the package (see Section \@ref(InstallPackages) for instructions) before running the function 


```{r loadDPlyR} 
library(dplyr) 
```  
 

﻿to gain access to the commands just listed. Note that the
`head()` command is used in many of the following examples to
ensure that only a handful of rows are printed out instead of the full
result. An alternative is to create a new data structure using the
`tbl\_df()` command. Let’s make a second copy of the air quality
data that is the same as the original except for being made ready for
use with the `dplyr` functions. 


```{r tbl_dfAirQuality} 
airquality2 = tbl_df(airquality) 
```  
 

﻿and then see the structure of the data using the `glimpse()` function that is a substitute for the `str()` command we’ve used previously. 


```{r GlimpseAirQuality} 
glimpse(airquality) 
```  
 

The `glimpse()` command works on a `data.frame` or the
new structure called a `tbl\_df` equally well.

The `filter()` and `arrange()` commands make many of the
examples seen earlier in this chapter much easier. The
`filter()` command is for extraction, while `arrange()`
is for rearrangement. For example, 


```{r GrabRows4} 
filter(airquality, Wind<10&Temp<60) 
```  
 

﻿extracts the rows of the `airquality` data that meet the
combinations of conditions specified. Use of `filter()` means we
don’t need the `data.frame` attached, or to gain access to the
variables in it using the dollar notation; we also don’t need the square
brackets for common subscripting tasks. We could have stored the outcome
in a new `data.frame` if we wished.

Perhaps the greatest gift of the `dplyr` package is its
popularising a different way of combining commands, called the pipe
operator. The last command issued can be re-written as: 


```{r GrabRows5} 
airquality %>% filter(Wind<10&Temp<60) 
```  
 

﻿This way of presenting commands is considered to be easier to read by
many users which helps with the popularity of the `dplyr` package. The pipe operator comes from another package called
`magrittr` so its use is not limited to the commands found in
the `dplyr` package. This way of presenting multiple commands
is called “chaining" in many resources. I like the simplicity that this
offers so I have tried to use the pipe operator as often as I can from
here on.

The `filter()` command is useful if we know the values a
variable takes, but at times we just want to know more about the
observations that are best or worst according to one (or more)
variables. For example, if we wanted to find the hottest days, we can
use the `arrange()` command instead of the `order()` command seen earlier. 


```{r HottestDays} 
airquality2 %>% arrange(desc(Temp)) 
```  
 

We can add more variables to sort the data by if we like. We used the
`desc` to get the temperatures to be in descending order in
this example.

The temperatures just listed are measured on the Fahrenheit scale. I
live in a country that uses the centigrade scale so a conversion is
needed to make these results more interpretable. Enter the
`mutate()` and `transmute()` commands. 


```{r MutateData} 
airquality %>% mutate(TempC = (Temp - 32) * 5 / 9) %>% head() 
```  
 

will create a new column in the returned data. N.B. the new data was not
stored in a new object so this command’s effect was purely temporary.

Again, notice that we didn’t need to do all that much to add a new
column and that this code is perhaps a little simpler than the
alternative: 


```{r MutateAlt, eval=FALSE} 
airquality$TempC = (airquality$Temp - 32) * 5 / 9 
```  
 

which we saw earlier. The main difference between the two commands is
that this one will add the new colun to the original
`data.frame`. In this instance the difference is probably
fairly trivial, but when we use much larger data sets, the storage of
results is important because of speed and memory usage concerns.

The command 


```{r TransmuteData} 
airquality2 %>% transmute(TempC = (Temp - 32) * 5 / 9) 
```  
 

﻿creates the new variable as a stand alone object. It hasn’t been stored
as an object so is only temporarily available.

Finally, the `dplyr` package also gives users two simple ways
to extract a random sample from a `data.frame`. The commands
`sample\_n()` and `sample\_frac()` extract a sample of a predetermined size and a specified fraction
respectively. For example: 


```{r ExtractSampleN} 
airquality %>% sample_n(4) 
```  
 

and 


```{r ExtractSampleFrac} 
airquality %>% sample_frac(0.05) 
```  
 

The `dplyr` package is under active development and so we
should expect more functionality to become available, but there is also
work being done to make the processing of data even faster so that it
can handle huge data sets. 

```{r cleanup, include=FALSE, purl=FALSE}
rm(list = ls())
``` 
 

