# LURN… To create maps{#Maps} 


```{r setup, include=FALSE, purl=FALSE} 
rm(list = ls())
if (!file.exists("Data")) dir.create("Data")
if (!file.exists("figures")) dir.create("figures")
if (!file.exists("tables")) dir.create("tables")
if (!file.exists("Other")) dir.create("Other")
set.seed(222664)
options(width = 60,  digits = 4, continue="   ")
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
Blind = file.exists("Blind.bat")
FigCap="To FIX"
``` 
 


```{r ChSetup, include=FALSE, purl=FALSE}
opts_chunk$set(fig.path='figures/') 
```  
 

This chapter explains how to create simple maps using some add-on
packages. You will need to have installed the `maps` ,
`mapdata` , and `ggmap` packages to complete the
examples in this chapter. If you have not installed an add-on package
before, you should take a look at Chapter \@ref(Additional) first.

After installing the packages, load them into your R session using: 


```{r MapsPackage} 
library(maps) 
library(mapdata) 
library(ggmap) 
```  
 

﻿## Creation of maps

The creation of maps is actually a simple task in R , as long as the
database for the desired land mass is available. There are a number of
these data bases available as downloadable packages from CRAN.

I first created a map when I wanted a map of New Zealand (my home
country) for a presentation I gave when I was overseas. I also wanted to
show a portion of the country to highlight how far my home town is from
the nation’s capital. The rectangle that appears on the map of New
Zealand was created using the `polygon()` command. The
`box()` command places the frame around the map which is created
using the `map()` command. Figure \@ref(fig:WholeNZ) shows the
first map, while Figure \@ref(fig:PartNZ) shows the central part of the
country.

```{r include=FALSE}
FigCap="Map of New Zealand with a rectangle marked on it that bounds the area used in the map presented in Figure \\@ref(fig:PartNZ)."
``` 
```{r WholeNZ, fig.cap=FigCap} 
map("nzHires", ylim=c(-48,-34), xlim=c(166,179), fill=T, col=3) 
box() 
polygon(x=c(172, 179, 179, 172), y=c(-42,-42, -39,-39), border=1, lty=2) 
```  
 

﻿[Link to SVG version](../SVGs/MapsWholeNZ-1.html)

The maps in these graphs are based on latitude and longitude data. The
`map()` function ensures the aspect ratio of the resulting plot
is in keeping with the actual view you would have from space if you were
directly above the centre of the map being plotted. Note that this rule
doesn’t apply for the world map options that exist.

## Adding cities to a map

The `map.cities()` function extracts data from a database of the
world’s cities. Judicious selection of the criteria for population was
required to get Palmerston North marked, and the `capitals` argument was used to get the point for Wellington. I have used some
colours to distinguish the different cities for my plot.

```{r include=FALSE}
FigCap="Map showing the central part of New Zealand in order to show two
cities in the lower North Island."
``` 

```{r PartNZ, fig.cap=FigCap} 
map("nzHires", ylim=c(-42, -39), xlim=c(172, 179), fill=TRUE, col=3) 
box() 
map.cities(minpop=70000,maxpop=80000, pch=16, col=2, cex=2) 
map.cities(capitals=1, col=4, pch=19) 
```  
 

﻿[Link to SVG version](../SVGs/MapsPartNZ-1.html)

One challenge in making this graphic presentable for this document was
assuring myself that the aspect ratio was correct. Latitudes and
longitudes do not take up equal space especially as you move further
from the tropics. R will get the aspect ratio correct but being sure is
another matter.

The graph I first created for this example was 5.7cm high and just over
10cm wide. Was it in proportion though? If each degree of latitude in
the map used 1cm of space, we would expect each degree of longitude to
take up at most 1 cm of space. An approximation for this is found by
multiplying the latitude space by $\cos(\theta)$ where $\theta$ is
the latitude. So each degree of longitude in my plot should take up
about `r round(cos(2*pi*40.5/360),2)`  of the space of
each degree of latitude. I’ve plotted three degrees of latitude and
seven degrees of longitude, so I expect my 5.7cm high graphic to be
$$\frac{5.7}
{3} \times 7 \times `r round(cos(2*pi*40.5/360),2)`  == `r round(1.9*7*cos(2*pi*40.5/360),1)` $$ centimetres in
width. This was about right.

You can see the way we would do this calculation using R in
Chapter \@ref(ScientificCalculator).

## Using Google’s services to map locations of interest


The database used for creating Figures  \@ref(fig:WholeNZ)  and \@ref(fig:PartNZ) 
predates the last New Zealand census. I now know that my home city no longer has a population in the range I used above. A smarter way to find locations is therefore necessary if the example is going to survive the next database update.


The ability to search Google for specific locations is really useful,
especially if you intend plotting maps with points of interest such as
universities rather than just cities. Our first example does look for
cities though, so that it builds on examples from previous sections in
this chapter.

In mid-2015, I attended the UseR conference in Aalborg, which is in
Denmark. I know Denmark has a coastline and is bordered by Germany, but
actually I don’t know much else (especially where individual cities
are\!). I wanted to know where the city of Aalborg is in relation to
Copenhagen (the capital of Denmark). 

I discovered that the functionality of this section is dependent on having a registration key with the Google Maps API. I could do this, but sharing my registration key is not smart. To run the rest of the example below, you will need to get your own registration key.

With the help of the
`ggmap` package tapping into Google maps, I can find the
latitute and longitude for both cities and then plot them on the map of
Denmark as follows: 


```{r include=FALSE}
FigCap="Map of Denmark using all data from the database."
``` 

```{r Denmark, fig.cap=FigCap} 
#Places = geocode(location=c("Aalborg", "Copenhagen")) 
#Places 
map('worldHires', 'Denmark') 
title('Denmark') 
#points(Places, pch=16) 
```  
 

```{r cleanup, include=FALSE, purl=FALSE}
rm(list = ls())
``` 
 

