# LURN… To Solve Operations Research Problems{#OperationsResearch} 


```{r setup, include=FALSE, purl=FALSE} 
rm(list = ls())
if (!file.exists("Data")) dir.create("Data")
if (!file.exists("figures")) dir.create("figures")
if (!file.exists("tables")) dir.create("tables")
if (!file.exists("Other")) dir.create("Other")
set.seed(222664)
options(width = 60,  digits = 4, continue="   ")
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
Blind = file.exists("Blind.bat")
FigCap="To FIX"
``` 
 


```{r ChSetup, include=FALSE, purl=FALSE}
opts_chunk$set(fig.path='figures/OperRes', out.width="0.7\\textwidth") 
require(lpSolve)
```  
 

﻿This chapter explains how to find solutions to several problems in
Operations Research. More specifically, they are problems in Linear
Programming (LP) and Integer Programming (IP) where the solution is
found as the optimal value of an objective function subject to the
existence of various constraints, all of which are expressed as linear
functions. The difference between LP and IP is the possible values for
the decision variables under consideration.

To benefit the most from this chapter, you ought to be familiar with
basic manipulation of vectors and matrices, as seen in
Chapter \@ref(LinearAlgebra) and be ready to install an additional package
using the methods outlined in Chapter \@ref(Additional). You will
need to install the `lpSolve` package and make sure it is
ready for use by calling the `require()` command as seen in the
examples in this chapter. Note that this need only be done once, but the
command is repeated as a reminder.

Note that all examples are intentionally made reproducible through use
of random number generation. You will need to make sure you reset the
random number generator’s seed value using the `set.seed()` commands that appear in the examples.

## The Assignment Problem

The assignment problem assigns *n* tasks to *n* people. The
objective function is often described in tabular or matrix form as a set
of costs to be minimised. The constraints are that all tasks must be
assigned and no person can complete more than one task. “Dummy" tasks or
people are added to the matrix if the number of tasks is not equal to
the number of people.

The mathematical programming representation for the assignment problem
has the objective function:
$$\min{ \sum_{i=1}
^n {\sum_{j=1}
^n {c_{ij}
x_{ij}
}}} \qquad i,j=1,\ldots,n$$

subject to the constraints: 
$$\sum_{j=1}
^n {x_{ij}} =1, \quad \forall i=1,\ldots,n$$
and$$ \sum_{i=1}
^n {x_{ij}} = 1, \quad \forall j=1,\ldots,n\\
x_{ij} &\in \{0,1\} \quad \forall i,j$$ 


```{r ORAssignment}
require(lpSolve)
set.seed(3010)
APCost.mat = matrix(sample(10:100, 16), nrow=4)
APCost.mat
Sol = lp.assign(APCost.mat)
names(Sol)
Sol$solution
Sol$objval
Sol
``` 
 

Note that the print out of the solution didn’t actually tell us which
task was to be completed by which person so we needed to look for the
matrix of the final values for the decision variables. There are plenty
of additional items stored as part of the solution, many of which you
may never need. It is assumed that if your needs run to the advanced
ideas using these items that you will now know where to find the
necessary information to solve your own problems.

## The transportation problem

The transportation problem aims to satisfy the demands of a set of
customers by optimising the selection of the supplier who will meet each
customer’s demands. The mathematical program for this problem has a cost
matrix and upper bounds on supply and lower bounds on demands. The
optimisation is a minimisation exercise so the demand is kept as low as
is required so the lower bounds are (in essence) equality constraints as
a consequence.

The objective function is:
$$\min{ \sum_{i=1}
^m{ \sum_{j=1}
^n{
c_{ij}
x_{ij}
}}}$$

and the constraints are: 
$$\sum_{j=1}
^n { x_{ij}
} \le s_i \quad \forall i=1,\ldots,m$$
 $$ \sum_{i=1}
^m { x_{ij}
} \ge d_j \quad \forall j=1,\ldots,n$$
and $$ x_{ij} \ge 0& \quad \forall i,j$$ 


```{r ORTransport}
require(lpSolve)
set.seed(3010)
TPCost.mat = matrix(sample(10:100, 30), nrow = 5)
TPCost.mat

DemandSigns = rep (">", 6)
DemandLimits = sample(50:300, 6)
DemandLimits
SupplySigns = rep ("<", 5)
SupplyLimits = sample(100:500, 5)
SupplyLimits
Sol = lp.transport (TPCost.mat, "min", SupplySigns, SupplyLimits, DemandSigns, DemandLimits, compute.sens=0)
Sol$solution
``` 
 

```{r cleanup, include=FALSE, purl=FALSE}
rm(list = ls())
``` 
 

