# LURN… To Analyze Quality Control Data{#QualityControl} 


```{r setup, include=FALSE, purl=FALSE} 
rm(list = ls())
if (!file.exists("Data")) dir.create("Data")
if (!file.exists("figures")) dir.create("figures")
if (!file.exists("tables")) dir.create("tables")
if (!file.exists("Other")) dir.create("Other")
set.seed(222664)
options(width = 60,  digits = 4, continue="   ")
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
Blind = file.exists("Blind.bat")
FigCap="To FIX"
``` 
 

```{r ChSetup, include=FALSE, purl=FALSE}
opts_chunk$set(fig.path='figures/QualityControl', tidy=FALSE, out.width="0.7\\textwidth") 
require(BrailleR)
```  
 

This chapter takes a basic look at the methods of quality control. This
material is not always taught in first year statistics courses and is
often one of those topics that some people think can be ignored. I guess
you can tell what I think given the inclusion of this topic in this
document. Like many other topics, you’ll need to consult a suitable
textbook for a thorough examination of the topic. Some would call
quality control or quality improvement a discipline not just a topic.

As quality control methods are not yet implemented in the base
distribution of R , you will need to obtain the `qcc` package
from CRAN, and install it. This can be done using the menus, command
lines, or using file management techniques from within your operating
system. Use the examples in Chapter \@ref(Additional) to help install
this package. Then use the `library()` command to make sure the
package is ready for action. 

```{r GetPackage} 
library(qcc) 
```  
 

I’ve used version `r packageVersion("qcc")`  of the
`qcc` package for the work in this chapter. Please note that
other packages have been developed that will also produce control charts
and other quality improvement analyses.

## Data

It is common for package developers to make data available to help users
test the functions of the package. The `qcc` package uses some
example from well-regarded textbooks as well as some other sources less
well-known. The context of the problems is of little value to the
presentation given here so I do not intend to use much time to explain
the data or the formulae needed to generate the details within each of
the various charts.

The one point that is very relevant to the presentation of the rest of
this chapter is the way that data are presented or stored. The
commonly-used `data.frame` with each column being a different
variable works well for many of the control charts covered here but is
not the easiest one to work with for the first few control charts we
investigate. The other way to present data is for each row to represent
a rational subgroup, and for columns to be used for elements within each
subgroup.. To get a mean of a subgroup we must use the
`rowMeans()` function for example.

## Monitoring a process mean

The most common control chart for monitoring a process mean is the x-bar
chart, often denoted $\bar{x}$ chart because the control statistics
being plotted is the sample mean of units within a rational subgroup of
data.

## Monitoring process variance

## Monitoring a process using proportions

## Monitoring a process using count data 

```{r } 
detach(package:qcc) 
```  
 

```{r cleanup, include=FALSE, purl=FALSE}
rm(list = ls())
``` 
 

