
# LURN… To Create Basic Graphs {#BasicGraphs} 


```{r setup, include=FALSE, purl=FALSE} 
rm(list = ls())
if (!file.exists("Data")) dir.create("Data")
if (!file.exists("figures")) dir.create("figures")
if (!file.exists("tables")) dir.create("tables")
if (!file.exists("Other")) dir.create("Other")
set.seed(222664)
options(width = 60,  digits = 4, continue="   ")
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
Blind = file.exists("Blind.bat")
FigCap="To FIX"
``` 
 

```{r ChSetup, include=FALSE, purl=FALSE}
opts_chunk$set(fig.path='figures/') 
require(BrailleR)
```  
 

﻿This chapter illustrates the construction of some basic exploratory data
analysis graphs. More complex graphs are considered in
Chapter \@ref(GraphsWithExtras), although at times we will see graphs
presented in conjunction with particular statistical analysis techniques
in other chapters. In this chapter, we concentrate on data taking
continuous values initially, but a short section on graphs for discrete
valued variables is also included.

There are many ways to create statistical graphs. The traditional method
uses one of the base R packages, but in recent years, the
`ggplot2` package [@Rpkg-ggplot2] has gained considerable attention. The
presentation of both styles of graph are given throughout this chapter.
You will need to ensure the `ggplot2` package is available for
use in your R session by issuing the command: 

```{r IntroAirQuality} 
str(airquality) 
```
  
 

It is common for R users to access the variables after issuing the
command 


```{r AttachAirQuality} 
attach(airquality) 
```  
 

﻿The `detach()` command undoes the `attach()` command. To remove the “attachment", you will issue the command
`detach(airquality)`.

You could look at the help file for this data if you wanted to learn its
complete story using `?airquality`. It tells us that the data
are for daily readings of the following air quality values for May 1,
1973 (a Tuesday) to September 30, 1973 (a Sunday).  - `Ozone`: Mean ozone in parts per billion from 1300 to    1500 hours at Roosevelt Island  - `Solar.R`: Solar radiation in Langleys in the frequency    band 4000–7700 Angstroms from 0800 to 1200 hours at Central Park  - `Wind`: Average wind speed in miles per hour at 0700 and    1000 hours at LaGuardia Airport  - `Temp`: Maximum daily temperature in degrees Fahrenheit    at La Guardia Airport.

This data suits our purposes for the majority of examples in this
chapter but we will also need to look at another data set for the
discussion of graphs for discrete valued variables.

## Histograms

Like many graphical functions in R , the `hist()` command will
attempt to make a suitably attractive histogram with the minimum of
input from the user. Figure \@ref(fig:AirQuality-HistWind) shows what
results from the simplest application of the `hist()` command.

Note that the figure created has default settings for the main title,
axis labels and that the number of classes (also called bins) and the
cutoffs between them have been chosen automatically. Various methods
exist for these choices, but it is my recommendation that the user find
out what happens when the default settings are chosen and then alter
only what is actually necessary. For example, graphs in this document do
not always need the default title inserted so we need to suppress the
default action if we want to remove the title. We may also want more
informative axis labels. Both of these alterations are done for the
creation of Figure \@ref(fig:AirQuality-HistWind2).

```{r include=FALSE}
BlindOrSightedText = ifelse(Blind, "The need to know what is printed in a histogram was a primary motivator for the development of the BrailleR package and is one of the examples given in Section \\@ref(VIExplained) on providing text descriptions of graphs. The graphical version of this histogram is described in text using the tools of the `BrailleR` package; you can learn how this is actually done later.",
"")
```

`r BlindOrSightedText` 

```{r include=FALSE}
FigCap="A histogram showing the wind speed measurements."
```

```{r include=FALSE}
SVGName = "AirQuality-HistWind"
```
```{r AirQuality-HistWind, eval=ifelse(Blind, -1, 1), echo=ifelse(Blind, -1, 1), fig.cap=FigCap} 
hist(Wind) 
MyHist = hist(Wind) 
    VI(MyHist)
```

```{r AirQuality-HistWindSVG, eval=Blind, echo=FALSE}
cat(paste0('
attach(airquality)
MyHist = hist(Wind) 
MakeAccessibleSVG(MyHist, "SVGs/', SVGName, '", view=FALSE)
detach(airquality)
'),
file="RunLater.R", append=TRUE)
```
[Link to SVG version](../SVGs/`r SVGName`.html)  

```{r include=FALSE}
BlindOrSightedText = ifelse(Blind, "### A note on the visual appearance of the standard histogram for blind users",
"")
```

`r BlindOrSightedText` 

`r if(Blind){Describe(MyHist)$general}` 

## Basic annotations to graphs

```{r include=FALSE}
FigCap="A histogram showing the wind speed measurements."
```
```{r include=FALSE}
SVGName = "AirQuality-HistWind2"
```
```{r AirQuality-HistWind2, eval=ifelse(Blind, -1, 1), echo=ifelse(Blind, -1, 1), fig.cap=FigCap} 
hist(Wind, xlab="Average wind speed (mph)", main="") 
MyHist = hist(Wind, xlab="Average wind speed (mph)", main="") 
    VI(MyHist)
```

```{r AirQuality-HistWindSVG2, eval=Blind, echo=FALSE}
cat(paste0('
attach(airquality)
MyHist = hist(Wind) 
MakeAccessibleSVG(MyHist, "SVGs/', SVGName, '", view=FALSE)
detach(airquality)
'),
file="RunLater.R", append=TRUE)
```
﻿[Link to SVG version](../SVGs/`r SVGName`.html)  

Variable names should be informative but aren’t always what we want to
see in graphs. Notice that in Figure \@ref(fig:AirQuality-HistWind2) ,
I’ve made the x-axis label more informative by indicating the units of
measurement for the wind speed. As I already have a caption for my
figure, I have chosen to remove the default title by adding the argument `main=""` to the `hist()`  command.

Aside from the alteration of the default axis label and the change in
the main title for our histograms, we could make quite a number of
changes. The `hist()` function allows the user various options
for the way the bars are filled in for example. It’s often worth
checking the default behaviour and then seeing if the resulting graph is
what you want. If it isn’t, then investigate your options by looking at
the relevant help file; in this case type `?hist` to get to
the help for the `hist()` function.

Other graphs we create will show points marked with small hollow
circles; we may wish to make these circles smaller, joined by line
segments, filled in, a different colour or combinations of these
attributes. The `par()` function should be investigated to find
out what is possible. Most graph functions allow the user the option of
adding arguments that will be passed directly to the `par()` function to obtain the same behaviour. Investigate the graphical
parameters at some stage using `?par` but be warned, there are
lots of adjustments that can be made. Experimenting is really the only
option when it comes time to get your graph looking perfect.

Additional text and/or lines can be added to some graph types. It may
prove useful to show the line of best fit along with the data
(illustrated in Chapter \@ref(Regression) ) for example. We’ll see how
to do the fancy things in Chapter \@ref(GraphsWithExtras).

## Other univariate summary graphs

Histograms aren’t the “be all and end all" for univariate summary
graphs. As a case in point, they aren’t at all useful for small sample
sizes. Various other graphs appear in introductory statistics courses
and that is why they appear here. I don’t mean to support one over
another at all — that’s up to you to determine.

### Boxplots

Boxplots show us quickly the shape of the distribution of a sample. They
show the median, lower and upper quartiles, and the minimum and maximum
of a sample. They will also identify points as outliers if these points
are too far from the bulk of values in the sample.

Figure \@ref(fig:AirQuality-BoxplotWind) shows us the boxplot for the average
wind speed at LaGuardia airport. Notice that I have added an additional
command to set up the size of the graph window. The `windows()` command has various aliases — `x11()` , `X11()` ,
`win.graph()` — none of which are required if the standard width
and height of the graph window are acceptable. You may wish to see why
I’ve changed the height of the graph window by ignoring the
`windows()` command from Figure \@ref(fig:AirQuality-BoxplotWind)

```{r include=FALSE}
FigCap="A boxplot showing the wind speed measurements."
```
```{r include=FALSE}
SVGName = "AirQuality-BoxplotWind"
```
```{r AirQuality-BoxplotWind, fig.width=7, fig.height=5, eval=ifelse(Blind, c(3,4), 2), echo=ifelse(Blind, c(1,3,4), c(1,2)), fig.cap=FigCap} 
windows(7, 5) 
boxplot(Wind, horizontal=TRUE, xlab="Wind speed (mph)")
MyBoxplot = boxplot(Wind, horizontal=TRUE, xlab="Wind speed (mph)")
VI(MyBoxplot)
```

```{r AirQuality-BoxplotWindSVG, eval=Blind, echo=FALSE}
cat(paste0('
attach(airquality)
MyBoxplot = boxplot(Wind, horizontal=TRUE, xlab="Wind speed (mph)")
MakeAccessibleSVG(MyBoxplot, "SVGs/', SVGName, '", view=FALSE)
detach(airquality)
'),
file="RunLater.R", append=TRUE)
```



 

﻿[Link to SVG version](../SVGs/`r SVGName`.html)  

Notice that I’ve changed the default orientation of the boxplot by
adding the argument `horizontal=TRUE` to the
`boxplot()` command. I have also ensured the more informative
axis label for wind speed is included using the `xlab` argument in the command.

### Comparative boxplots

Boxplots are often useful for comparing several small samples at once.
We must ensure that the same axis is used for the units of measure of
interest and the easiest way to ensure this is to put the various
samples into the same graphic with only one axis rather than having a
series of single boxplots each having their own axis.

For the purposes of illustration, I want to show the distribution of the
daily average wind speeds for the five months separately.

```{r include=FALSE}
FigCap="A comparative boxplot showing the wind speed measurements for each month."
```
```{r include=FALSE}
SVGName = "AirQuality-CompBoxplotWindMonth"
```
```{r AirQuality-CompBoxplotWindMonth, fig.width=7, fig.height=5, eval=ifelse(Blind, c(3,4), 2), echo=ifelse(Blind, c(1,3,4), c(1,2)), fig.cap=FigCap} 
windows(7, 5) 
boxplot(Wind~Month, xlab="Month", ylab="Wind speed (mph)")
MyBoxplot = boxplot(Wind~Month, xlab="Month", ylab="Wind speed (mph)")
VI(MyBoxplot)
```

```{r AirQuality-CompBoxplotWindMonthSVG, eval=Blind, echo=FALSE}
cat(paste0('
attach(airquality)
MyBoxplot = boxplot(Wind, horizontal=TRUE, xlab="Wind speed (mph)")
MakeAccessibleSVG(MyBoxplot, "SVGs/', SVGName, '", view=FALSE)
detach(airquality)
'),
file="RunLater.R", append=TRUE)
```
﻿[Link to SVG version](../SVGs/`r SVGName`.html)  

This comparative boxplot is created using a formula to describe the
relationship between the two variables that are referred to in our
graph. The use of the tilde symbol between `Wind` and `Month`  could be read as something like “average wind speed
depends on the month" — well a theory that might be illustrated in our
graph anyway. Certainly it is the potential for this relationship to
exist that may be exposed through use of the comparative boxplot.

### Dotplots

Dotplots aren’t everyone’s cup of tea but they are frequently offered as
substitutes for boxplots and histograms. Again, I chose to alter the
default window size for the example given in
Figure \@ref(fig:AirQuality-DotPlotWind) because I didn’t like the way the regular graph window presented this graph.


```{r include=FALSE}
FigCap="A dotplot showing the wind speed measurements."
```
```{r include=FALSE}
SVGName = "AirQuality-DotplotWind"
```
```{r AirQuality-DotPlotWind, fig.width=7, fig.height=5, eval=ifelse(Blind, c(3,4), 2), echo=ifelse(Blind, c(1,3,4), c(1,2)), fig.cap=FigCap} 
windows(7, 2.5) 
dotchart(Wind) 
MyDotplot = dotchart(Wind) 
VI(MyDotplot)
```

```{r AirQuality-DotplotWindSVG, eval=Blind, echo=FALSE}
cat(paste0('
attach(airquality)
MyDotplot = dotchart(Wind) 
MakeAccessibleSVG(MyDotplot, "SVGs/', SVGName, '", view=FALSE)
detach(airquality)
'),
file="RunLater.R", append=TRUE)
```
﻿[Link to SVG version](../SVGs/`r SVGName`.html)  

Some people find the standard dotplot that R creates a little
artificial. The spacing between points horizontally and vertically is
captured by the human eye, but this graph is a single dimensional
representation. Adding jitter to the data allows points where there are
ties to be represented by a pair of points on the graph rather than two
perfectly overlaid points which look like a single point. The
`jitter()` command can be embedded within the command for
creation of a dotplot, but is not required for our example data as there
are no ties. If there were a large number of ties, we would have used
the command `dotchart(jitter(Wind))`.

### Simple line plots {#LinePlots}

Occasionally it’s useful to see how a measurement changes over the time
the data were collected. R does this very simply using the
`plot()` command as shown in
Figure \@ref(fig:AirQuality-LinePlotTemp) which is for the maximum daily
air temperatures at LaGuardia airport from 1 May to 30 September 1973.
We can see periods of time where the maximum temperatures were fairly
consistent and periods where it was fairly volatile. The middle of the
graph is for the month of July which is the height of summer in New York
and as a consequence we expect to see few points nearer the lower part
of the graph during this period.


```{r include=FALSE}
FigCap="A plot showing the temperature measurements in the order they were collected."
```
```{r AirQuality-LinePlotTemp, fig.cap=FigCap}
plot(Temp, ylim=c(50, 100)) 
``` 

 

```{r AirQuality-LinePlotTemp2, fig.cap=FigCap} 
plot(Temp, ylim=c(50, 100), type="l") 
``` 
 

﻿Notice that we have altered the range of values covered by the
*y*-axis using a specific command. The `ylim` has a
corresponding `xlim`  to create limits for the *x*-axis.
Adding one more argument to the `plot()` command will change the plotting from points to lines (shown in
Figure \@ref(fig:AirQuality-LinePlotTemp2) ). Combinations of points and
lines can be obtained (not shown); the user can also alter the style of
the points and lines being printed.

There is a simpler way to generate time series plots which we
demonstrate in Chapter \@ref(TimeSeries). It is easier to augment
this line plot than the time series plot, and in so doing we will gain
an insight into how other plots like the time series plot are created.

## Quantile-quantile plots

The most common quantile-quantile plot we might wish to create is used
to investigate the usefulness of the normal distribution to model a
variable’s distribution. Normal quantiles are created automatically for
the normal quantile plot when it is generated using the
`qqnorm()` command. The default plot for this is shown in
Figure \@ref(fig:AirQuality-NormPlotWind).

If these data were normally distributed, the points on the plot would
lie on a straight line. The `qqline()` command adds the straight
line to the plot to assist with determining the linearity of the points.

```{r include=FALSE}
FigCap="A quantile-quantile plot showing the normality of the wind speed measurements."
```
```{r AirQuality-NormPlotWind, fig.cap=FigCap} 
qqnorm(Wind) 
qqline(Wind) 
``` 
 

﻿
```{r include=FALSE}
BlindOrSightedText = ifelse(Blind, "There are numerous formal hypothesis tests that can assess the normality of a variable. Some of them are discussed in Section \\@ref(NormalityTests); they are  fairly easy to apply.",
"")
```

`r BlindOrSightedText` ## Scatter plots {#ScatterPlots}

Scatter plots are created using the `plot()` command by one of
two methods. Figure \@ref(fig:AirQuality-ScatterTempWind) was created
using 

```{r ScatterCode1, eval=FALSE} 
plot(Wind, Temp) 
```  
 

but the same plot can be generated using what is known as a formula. In
this case, only one argument is given to the `plot()`  command
but both variables of interest are in that argument. 

```{r ScatterCode2, eval=FALSE} 
plot(Temp~Wind) 
```  
 

﻿The tilde symbol is often read as “…is distributed as…" but we might
simplify this to be read as “follows". This makes some sense as we
generally create a scatter plot to see if one variable follows the
other; in this case we are seeing if temperature depends on wind speed
in some way. I prefer this approach as it matches what we do when fitting a regression model, see Chapter \@ref(Regression) to see this demonstrated.


```{r include=FALSE}
FigCap = "A scatter plot showing temperature on the y axis and wind speed on the x axis."
```
```{r AirQuality-ScatterTempWind, fig.cap=FigCap} 
plot(Wind, Temp) 
``` 
 


## Scatter plot matrices {#ScatterPlotMatrices}

A scatter plot matrix is just a matrix of scatter plots where each
variable is plotted against all other variables. This graphic is
therefore very useful for a preliminary look at multivariate data. In R
, we obtain this graphic using the `pairs()` command as
illustrated in Figure \@ref(fig:AirQuality-ScatterPlotMatrix).

Only four of the variables within the `air quality` data have
been used in this example because the variables for the month and day
take discrete values and therefore do not suit scatter plots — take a
look for yourself if you must but it’s probably better to think about
why this is the case before you look at the graphs. To select the four
variables of interest, I have created a `data.frame` with the
variables I want included; this `data.frame()` command is then
nested within the `pairs()` command.


```{r include=FALSE}
FigCap = "A scatter plot matrix for four variables found in the the Air Quality data."
```
```{r AirQuality-ScatterPlotMatrix, fig.cap=FigCap} 
pairs(data.frame(Ozone, Solar.R, Wind, Temp)) 
```  
 

Note that the names of the variables appear in the spaces on the leading
diagonal and that the graphs on either side of the diagonal plot the
same data but with the axes reversed. Sometimes the human eye will pick
up a relationship when the data are presented one way better than the
other way.

## Graphs for discrete valued variables

R does not contain many graphs for discrete valued variables.

### Bar charts {#BarCharts}

If a variable is considered by R to be a `factor` , then the
default action of the `plot()` command is to construct a bar
chart. This is illustrated using a data set which is part of the default
installation of R called `state.region` examined using 

```{r IntroStateRegion} 
str(state.region) 
levels(state.region) 
```  
 

﻿Given this variable is a `factor` with four levels it is well
suited to presentation in a bar chart, as seen in
Figure \@ref(fig:StateRegion-BarChart).

```{r include=FALSE}
FigCap="A bar chart showing which of the regions each of the fifty U.S.
states belongs"
``` 

```{r StateRegion-BarChart, fig.cap=FigCap} 
plot(state.region) 
```  
 

This figure was created from the raw data, that is the regions for each
of the fifty states in the U.S. If we had summary data with counts for
each of the categories, we would need to use the `barplot()` command. The `summary()` command shows us how many states fall
into each category in this instance. 

```{r SummaryState} 
summary(state.region) 
```  
 

﻿These values are then plotted in Figure \@ref(fig:StateRegion-BarChart2).

```{r include=FALSE}
FigCap="A bar chart showing which of the regions each of the fifty U.S.
states belongs"
```
```{r StateRegion-BarChart2, fig.cap=FigCap} 
barplot(summary(state.region)) 
```  
 

﻿While this data set is rather trivial, it is useful for demonstrating
one more feature. Note that both in the output above and in
Figure \@ref(fig:StateRegion-BarChart) , the Western region is the last
category. To reorder the regions in the bar chart is actually fairly
straight forward. All we need to do is make a small addition to the
existing commands. 

```{r ReorderRegions} 
summary(state.region)[c(4,2,3,1)] 
```  
 

﻿and re-create the bar chart accordingly (see
Figure \@ref(fig:StateRegion-BarChartImp) ).

```{r include=FALSE}
FigCap="An improved bar chart showing which of the regions each of the
fifty U.S. states belongs"
``` 
```{r StateRegion-BarChartImp, fig.cap=FigCap} 
barplot(summary(state.region)[c(4,2,3,1)]) 
```  
 

﻿
### Pie charts {#PieCharts}

OK, if you must do so, make a pie chart using the `pie()` command. Even the help for this command says they are a bad
representation for data, stating “Pie charts are a very bad way of
displaying information. The eye is good at judging linear measures and
bad at judging relative areas. A bar chart or dot chart is a preferable
way of displaying this type of data."

The pie chart for the `state.region` data is given in
Figure (fig:StateRegion-PieChart).

```{r include=FALSE}
FigCap="A pie chart showing which of the regions each of the fifty U.S.
states belongs"
```
```{r StateRegion-PieChart, fig.width=7, fig.height=7, fig.cap=FigCap} 
pie(summary(state.region)) 
```  
 

## Closing

If you are carrying on working with R you might wish to remove direct
access to the data sets we used in this chapter by issuing the following
commands 

```{r DetachAirQuality} 
detach(airquality) 
``` 
 

```{r cleanup, include=FALSE, purl=FALSE}
rm(list = ls())
``` 
 

