# LURN… To Perform Times Series Analyses{#TimeSeries} 


```{r setup, include=FALSE, purl=FALSE} 
rm(list = ls())
if (!file.exists("Data")) dir.create("Data")
if (!file.exists("figures")) dir.create("figures")
if (!file.exists("tables")) dir.create("tables")
if (!file.exists("Other")) dir.create("Other")
set.seed(222664)
options(width = 60,  digits = 4, continue="   ")
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
Blind = file.exists("Blind.bat")
FigCap="To FIX"
``` 
 


```{r ChSetup, include=FALSE, purl=FALSE}
opts_chunk$set(fig.path='figures/') 
chooseCRANmirror(ind=1)
if(!require(tseries)) install.packages("tseries")
if(!require(forecast)) install.packages("forecast")
library(forecast)
library(tseries)
library(BrailleR)
```  
 

This chapter focuses on some simple time series approaches for
univariate time series. At this stage multivariate time series are not
included, but could be in future depending on demand. In any case, the
methods shown here are certainly not meant to be a comprehensive
elucidation of all time series methods. Dealing with just the most
commonly taught (and therefore the simplest) methods takes up enough
space.

You will need to refer to a standard time series analysis textbook for
more complete guidance; many of these are now written with R as the
chosen software to support the material in the book.

Given this proliferation of textbooks on the topic of time series
analysis, it is no surprise that there are a number of add-on packages
written to support them. In this chapter we only extend beyond the
functionality offered by the base distribution of R with two extra packages.

## Time series objects in R 

The basic univariate time series, with no missing data is stored in R as
a time series object. This is marginally different to a vector in that
it has a time stamp or index associated with each observation in the
series.

We will use several time series data sets that are available within the `datasets` package [@Rpkg-base]  which is part of the default installation of R. 


```{r IntroData} 
str(LakeHuron) 
str(lynx) 
str(Nile) 
```  
 

The first is for the level of Lake Huron; the second is the number of
lynx that were trapped in Canada; and the third is the flow of the River
Nile. More detailed information for all of these series is available
using the relevant help function for the series.

You can use the `as.ts()` command to convert a vector of
observations to a time series object. For example, the
`airquality`data is recorded in time order. We might
want to treat the observed temperatures as a time series by converting
it using: 


```{r MakeTSTemp} 
Temp.ts = as.ts(airquality$Temp) 
```  
 

## Time series plots

We have seen the flexibility of the `plot()` command in previous
chapters, and now we see another use. Applying the `plot()` command to a time series object generates a time series plot, as seen in
Figure \@ref(fig:NileTSPlot).

```{r include=FALSE}
FigCap="Time series plot for the annual flow of the River Nile."
``` 
```{r NileTSPlot, fig.cap=FigCap} 
plot(Nile, xlab="Year", ylab="Flow") 
``` 

```{r NilePlotSVG} 
BrailleR::TimeSeriesPlot(Nile, xlab="Year", ylab="Flow") 
``` 
 

﻿[Link to SVG version](../SVGs/TimeSeriesNilePlot-1.html)

Having the class assigned to our data means the plot was enhanced
slightly. Check this by comparing the results for the two plots in
Figure \@ref(fig:TSPlotsCompared).

```{r include=FALSE}
FigCap="Comparison of use of plot on a vector and a time series
object."
``` 
```{r TSPlotsCompared, fig.height=8, fig.cap=FigCap} 
par(mfrow=c(2,1))
plot(airquality$Temp, ylab="Temperature") 
plot(Temp.ts, ylab="Temperature") 
```  
 

﻿[Link to SVG version](../SVGs/TSPlotsCompared-1.html)

[Link to SVG version](../SVGs/TSPlotsCompared-2.html)
```{r include=FALSE}
BlindOrSightedText = ifelse(Blind, "### A note on the visual appearance of the standard time series plot (written for blind users)",
"")
```

`r BlindOrSightedText` `r if(Blind){Describe(TimeSeriesPlot(airquality$Wind))$general}` 

## Smoothing of a time series using moving averages {#MovingAverages}

One simple method for smoothing out a time series in order to expose its
behaviour is to smooth out the observed data using the
`filter()` command to create a moving average. A moving average
is often centred, (using the argument `sides=2` ) but if the
series is expected to have no trend, then a backwards only
(`sides=1` ) focus can be taken. We must determine the width
of the moving average, usually using odd-numbered widths.

Figure \@ref(fig:FilterExample) shows how different amounts of data can
be averaged to change the amount of smoothing.

```{r include=FALSE}
FigCap="Comparison of different amounts of smoothing a time
series."
```

```{r FilterExample, fig.cap=FigCap} 
plot(Nile) 
for(i in 1:4){ 
w=4*i-1 # choose the window (3, 7,11,15) 
lines( (start(Nile)[1]):end(Nile)[1], filter(Nile, rep(1/w, w), sides=1), col=i+2) 
} 
```  
 

[Link to SVG version](../SVGs/FilterExample-1.html)

## Checking for stationarity {#Stationarity}

Stationarity is important for modelling time series. Some descriptive methods also
need it. The simplest form of stationarity is when the mean of the
series and the variance of the series both remain roughly constant over
time, or as explained by Hyndman and Athana-sopou-los (2013), "A stationary time series is one whose properties do not depend on the time
at which the series is observed". (their book is called Forecasting:
Principles and Practice.)

A white noise series of normally distributed data with mean zero and
constant variance $\sigma^2$ is the strongest form of stationarity we
seek in our modelling. It is the desired outcome for the residuals from
any model we fit to time series data.

The `Box.test()` function tests for non-stationarity. 


```{r BoxTest} 
Box.test(Nile, lag=20) 
```  
 

while several other tests exist in add-on packages which can be accessed
by loading the `tseries` and `forecast` packages. 


```{r ForecastTests} 
library(tseries) 
library(forecast) 
adf.test(Nile) 
pp.test(Nile) 
kpss.test(Nile) 
```  
 

﻿Each test is looking at a different facet of what might indicate how a
series might not be stationary. Check out the null hypothesis and
alternative for each test on its help page.

## Autocorrelation and partial autocorrelation

The autocorrelation and partial autocorrelation function values can be
obtained using the `acf()` and `pacf()` commands.
Plotting these is a common way to determine if any of the values found
are of interest in understanding the process being modelled so this is
the default action performed by these commands. See
Figure \@ref(fig:NileACF) for the plot generated by the `acf()` command. The `pacf()` command functions in the same way so is
not demonstrated.

```{r include=FALSE}
FigCap="Autocorrelation function for the annual flow of the River
Nile."
``` 
```{r NileACF, fig.cap=FigCap} 
acf(Nile) 
```  
 

[Link to SVG version](../SVGs/TimeSeriesNileACF-1.html)

To obtain the values of the autocorrelation function (or partial
autocorrelation function) in text form, we add an argument to stop the
results being plotted. 


```{r NileACFNoPlot} 
acf(Nile, plot=FALSE) 
pacf(Nile, plot=FALSE) 
```  
 

## Decomposition into seasonal and trend components

The basic idea of the decompositions presented in this section is to
explain the observed time series values $y_t$ in terms of a trend
$T_t$, a seasonal component $S_t$ and residual error.

Obviously we need a series that has a seasonal component such as the
Australian resident numbers measured quarterly from 1971 to 1994. 


```{r getAustResData} 
str(austres) 
```  
 

The `decompose()` function can create two forms of the
decomposition, one of which is additive and the other is multiplicative. 


```{r DecomposeAustRes} 
AustRes.dec1 = decompose(austres) 
Aust.dec2 =decompose(austres, type="mul") 
```  
 

The `stl()` command uses loess smoothing to estimate the
trend component before finding the seasonal components and the resulting
error terms, and is a slightly more advanced process than that offered
by `decompose()`. N.B. there is an `stl()` command in
the `stats` package and another in the `forecast` package. We use the simpler one here but need to force R to do so just
in case the `forecast` package version has precedence. 


```{r STLAustRes, fig.cap=FigCap} 
AustRes.dec3 = stats::stl(austres, s.window="periodic") 
AustRes.dec3 
plot(AustRes.dec3) 
```  
 

## Exponential smoothing {#ExponentialSmoothing}

Exponential smoothing is a method for making predictions of the next
observation in a time series. The next observation is a weighted average
of all observations to date, with the most recent given the greatest
weight and the oldest ones having the least impact. This simple version
is useful for situations with no seasonal component or systematic trend
component.

Note that this is often expressed as $$\hat{y}
_{t+1} = \lambda \hat{y}
_{t} + (1-\lambda)y_{t}$$ Substituting for $\hat{y}
_t$ in terms of terms at time \*\*\*t-1\*\*\* shows how older values
from the observed series contribute smaller and smaller amounts to the
new prediction.

The `HoltWinters()` command is used for a wide range of models
that include the exponential smoother above and allows for trend and
seasonality components to be introduced. Starting with the Nile data: 


```{r NileHoltWinters, fig.cap=FigCap} 
Nile.hw1 = HoltWinters(Nile, beta=FALSE, gamma = FALSE) 
Nile.hw1 
plot(Nile.hw1) 
```    
 

Note that we must explicitly set `beta` and `gamma` to `FALSE` to get the simple exponential smoother.

Adding a linear trend is fairly simple, but the seasonal component can
be expressed as either an additive or a multiplicative term. The two
forms would be $$y_t = \mu_t + \beta_t t +S_{t-p} + e_t$$ and $$y_t = (\mu_t + \beta_t t) S_{t-p} + e_t$$ respectively , where the $\mu_t$ is found using a
deseasonalised version of the simple exponential smoother
$$\mu_t = \alpha (y_t - S_{t-p}
) + (1-\alpha) (\mu_{t-1} + \beta_{t-1}
)$$ for the additive model and $$\mu_t = \alpha \frac{y_t }
{ S_{t-p}
} + (1-\alpha) (\mu_{t-1} + \beta_{t-1}
)$$ for the multiplicative model. 


```{r AustResHW, fig.cap=FigCap} 
AustRes.hw1 = HoltWinters(austres) 
AustRes.hw1 
plot(AustRes.hw1) 
AustRes.hw2 = HoltWinters(austres, seasonal="mult") 
AustRes.hw2 
plot(AustRes.hw2) 
```  
 

﻿## Autoregressive models

ar()

## Basic ARIMA models

ARIMA models require the autoregressive and moving average components to
be built on a stationary series. We often need to create a stationary
series using differencing as seen in Section \@ref(Stationarity)
above. arima()

## Seasonality and ARIMA modelling 

```{r cleanup, include=FALSE, purl=FALSE}
rm(list = ls())
``` 
 

