# LURN… The basics of the BrailleR add-on package{#BrailleR} 


```{r setup, include=FALSE, purl=FALSE} 
rm(list = ls())
if (!file.exists("Data")) dir.create("Data")
if (!file.exists("figures")) dir.create("figures")
if (!file.exists("tables")) dir.create("tables")
if (!file.exists("Other")) dir.create("Other")
set.seed(222664)
options(width = 60,  digits = 4, continue="   ")
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
Blind = file.exists("Blind.bat")
FigCap="To FIX"
``` 
 


```{r ChSetup, include=FALSE, purl=FALSE}
opts_chunk$set(fig.path='figures/VI', out.width="0.7\\textwidth") 
```  
 

﻿This chapter provides a brief explanation of how the
`BrailleR` package [@Rpkg-BrailleR] can be used by blind people to get
many tasks done quickly and effectively. In general, the package does
three things:

1.  creates a copy of the R console/terminal in a text file including    all standard output.

2.  creates text descriptions of some graphs. These tools should help    the blind user create graphs (with confidence) for the sighted world    we must work in.

3.  provides convenience functions for novice users who do not know    enough R commands to survive. Sighted people have the benefit of    graphical user interfaces (GUI) to help them, but none of them are    accessible by screen reading software.

Initially, the package was built on top of functionality I wrote to
support my own work practices, but additions to meet the needs of other
blind R users have been made on request.

A much more comprehensive user’s guide to the `BrailleR` package can be found at the [BrailleR in Action e-book homepage.](https://R-Resources.massey.ac.nz)

You will need to install the `BrailleR` package before
continuing on with the examples in this chapter. It is available from
CRAN. Use the examples in Chapter \@ref(Additional) to help install
this package. Then use the `library()` command to make sure the
package is ready for action. 


```{r LoadPackage} 
library(BrailleR) 
```  
 

﻿This chapter was created using version `r packageVersion("BrailleR")`  of the `BrailleR` package. This version of the package has a number of other packages that
it depends on for functionality. When you make the `BrailleR` package available, you may see a few introduction messages that tell you
about the other packages being loaded. You will also be told about some
default settings for the package that can be altered later to suit your
needs.

If you want to know more about the `BrailleR` package and its
creation, you might like to look over its homepage by issuing the
`BrailleRHome()` command which will take you to the [BrailleR Project homepage.](https://R-Resources.massey.ac.nz/BrailleR)

Once you have read through this chapter, you can check out the rest of
the `BrailleR` package’s features as explained in another
e-book. The title of that e-book is “BrailleR in Action"; you can go to
its homepage from R directly by issuing the `BrailleRInAction()` command which will take you to the front page of the book at
https://R-Resources.massey.ac.nz/BrailleRInAction

## Creating a copy of the R console window {#CreateCopyConsole}

In Chapter \@ref(BlindUse) I mentioned the need to save the content of
the console window to a text file to be able to copy and paste its
content to a report or document. One major feature of the `BrailleR` package is that this functionality can be done easily and efficiently.

There are three types of content printed in the console window.

1.  The commands we type,

2.  The output valid commands generate, and

3.  Any error or warning messages that arise from our commands.

The `sink()` command does not currently record the commands or
the errors and warning messages. It is therefore a valuable tool for
those R users that are confident about the validity of their scripts.

Similarly, the `history()`  command shows a list of all issued
commands, even those that contain errors. N.B. The `BrailleR` package has its own `history()` command that opens the contents
of the command history in a text editor so that it can be read using
screen-reading software.

Saving the commands typed into an R session can be achieved using the
`savehistory()` command, but again, this will include the
commands that generated errors. For example 


```{r SaveHistory, eval=FALSE} 
savehistory("WhatIDidToday.txt") 
```  
 

will save all the issued commands into a text file called `
WhatIDidToday.txt` in the current working directory.

If I just want to have everything I do during a session recorded for
posterity, I want all contents of the console window to be saved. A
plain text file is sufficient but it is possible to use add-on packages
to create other file formats.

I have included a convenience function in the `BrailleR` package that starts a text file with the date and time in its filename.
It gets saved in the current working directory. To get this file
started, just type 


```{r TXTOut, eval=FALSE} 
txtOut() 
```  
 

﻿In fact this command starts several files, one for the complete console
and one for the issued commands. The command actually runs the
`txtStart()` command but specifies particular options for the
user’s convenience. If you are happy to use the default filename then you
just need to hit `<enter>`, but if you want to specify the filename, you
type the name, and then hit `<Enter>` twice (once for your file and once
to get out of the filename requesting prompt).

See the help page for the `txtOut()` command as it includes the
history of the command and its relatives. The other related commands add
comments to the text file while `txtStop()` stops the processing
entirely. Also note that the help page mentions how to get the packages
needed for other file formats.

You can open the text file that is being created and modified as you
work in any browser. Once you have the file open and have done some
work, you must refresh your browser to see the latest changes. Keep the
browser open as you work and remember to refresh often to keep on top of
what is happening with complete access via your screen reader.

## Text interpretation of graphs {#VIExplained}

Statistics is an inherently visual discipline. Many statistical analyses
lead to a graphic representation of data that is easy for the sighted
world to interpret. Through use of R we can create functions that can
interpret the graphs using text descriptions. We’ll start with a
histogram.

In most circumstances, the user will create a histogram using the
`hist()` command. This command takes data and creates a number
of values needed to construct the histogram. We can see what has been
created by storing these values in an object. For example: 

```{r CreateHist, warnings=FALSE} 
MyHist = hist(airquality$Wind, xlab="Average wind speed (mph)", main="", plot=FALSE) 
```  
 

The `plot=FALSE` argument stops the graph being plotted. We can get
the graph anytime by issuing the `plot()` command but we are
only interested in seeing what R has stored in the `MyHist` object at the moment. 


```{r ShowPrintHist} 
MyHist 
str(MyHist) 
class(MyHist) 
```  
 

﻿The full print out has everything we need but is not in a friendly
format. The `str()` command gives us a condensed version of the
data being stored but isn’t sufficient, and the `class()` command just tells us that this information is stored in an object of
type `histogram`. The class of the object is the most
important as it is what tells R how to work on the object when we use
the `plot()` command on the `MyHist` object. The `plot()` command is actually a family of commands, one of which
is purposely designed for histograms. The `plot.histogram()` command is actually what constructs the graph when the user calls the
`plot()` command on an object of type `histogram`.
This family of commands is known as a "method".

The `BrailleR` package includes a method for creating text
interpretations in a similar way. The `VI()` command is actually
a family of commands that includes the `VI.histogram()` command.
When we issue the `VI()` command on the `MyHist` object, the `VI.histogram()` function does the work. Let’s see what we get: 

```{r Hist, fig.cap=FigCap}
VI(hist(airquality$Wind, xlab="Average wind speed (mph)", main="")) 
```

```{r HistSVG, eval=Blind, echo=FALSE, include=FALSE, messages=FALSE, warnings=FALSE, results="hide"} 
MakeAccessibleSVG(hist(airquality$Wind, xlab="Average wind speed (mph)", main=""), "figures/VIHistAirQualityWind") 
``` 
[Link to SVG version](../SVGs/VIHistAirQualityWind.html)  
 

Other functions exist for boxplots and dotplots. Try: 


```{r VIBoxplot} 
VI(boxplot(airquality$Wind, xlab="Average wind speed (mph)")) 
```  
 

and 

```{r VIDotplot} 
VI(dotplot(airquality$Wind, xlab="Average wind speed (mph)")) 
``` 
 

﻿The `dotplot()` command is not part of base R which uses the
`stripchart()` command instead. The `hist()` and `boxplot()`  commands used here are actually commands from the
`BrailleR` package that use the standard commands indirectly.
At this point in time, I haven’t worked out how to get the
`stripchart()` command to work as a `BrailleR` command so I needed to use a different name for my version. The
`BrailleR` versions of these three commands store the
additional information we need to get the text interpretation that is
not possible from the standard version of the commands.

## Basic descriptions of variables

The `UniDesc()` function is designed to quickly generate a text
summary of a single variable. It also saves several graphs. It can also
save the text in a file and use the `VI()` command from the
`BrailleR` package to add text descriptions to the output
file.

One obvious issue is that saving lots of files could become problematic.
Default filenames are used and all files can be put into a folder of the
user’s choosing.

An example will help. Try 


```{r UniDesc, eval=FALSE} 
data(airquality) 
Ozone=airquality$Ozone 
UniDesc(Ozone, Folder="AirQuality") 
```  
 

﻿Take a look in the `AirQuality` folder that has been created in your working
directory. It has a number of files that all start with “Ozone" --- the
name of the variable being analysed. You will also have three files
stored in the current working directory which have the filename `Ozone-UniDesc.*` where
the star is for the extension which is one of `html`, `R`, and `Rmd`.

The most useful one at first is the `Ozone-UniDesc.html` file. Open this file in a browser to
see what analyses have been included. This file is specifically
formatted for use with screen readers. I hope you find it valuable. If
you are reading this document on the web, then the link
[Ozone-UniDesc.html](Ozone-UniDesc.html) will open it in your browser. N.B. This link will not work if you are using the \file{pdf} version of LURN.

The commands used to generate parts of this HTML document are given in
the R script file; the third file is called an R markdown document. This
is the file that gets converted into the other two files. It might prove
useful to learn how to write this kind of file for yourself one day.

In the subfolder you will find that graphs have been saved in a variety
of formats. They each have their uses and hopefully the filetype you
need is there. Special attention is made for those users of LaTeX who
need specific formats for graphics. Some more advanced information is
presented in LaTeX formatted tables and put into files with the `.tex` extension. 

Whenever the `VI()` function is employed on an object that does
not yet have an explicit function tied to it, then you will be told that
this is the case. This package is a work in progress after all. Please
feel free to send an email if you’d like any extra functionality.

## Altering R output to make it easier to read

Some R output is arranged in nice tables that are easy for the sighted
user to read, but are difficult to understand if using synthesised
speech to read the information back to you. The `VI()` command
is used again here but is now employed on a `data.frame` object. We’ll compare it with the `summary()` command used frequently. 


```{r VIDataFrame} 
summary(airquality) 
VI(airquality) 
```  
 

﻿The output generated by the `summary()` command is difficult to
follow as variables are represented in columns while rows that are read
aloud by a screen reader are for the sample statistics.

The `VI()` method has given us a summary of the variables one by
one. This should prove easier to navigate and interpret as either a
braille or screen reader user.

## Reading a scatter plot

One challenge facing blind people is the inability to glance at a graph
and understand what is intended just as sighted people can. The most
difficult graphs to deal with are multidimensional representations of
data. Two dimensional scatter plots are the first example of such
graphs.

When a sighted person looks at a scatter plot, they are looking for a
number of things. A relationship between the variables is the most
common thing to look for, but there is also a need to identify points
that are unusual in the context of the data presented.

One possible way to describe how the data points are summarised at a
glance is that the graphing window is broken up into areas and a rough
guess at the density of points within each region is evaluated. If this
evaluation of density can be guessed, it can be counted more exactly by
a purposely written function. The `WhereXY()` function does this
counting for us. It assumes equal-sized rectangular regions in the graph
window. We can choose how many regions but the illustrations here all
use a $5\times5$ grid.

Let’s work with a number of examples, all based on a set of
predetermined *x*-values and various possible *y*-values. 


```{r CreateVars} 
x=(0:100)/100 
y1=5*x+rnorm(101) 
y2=(2*x-1)^2+rnorm(101)/6 
y3=11*x+rnorm(101) 
y4=y3; y4[100]=0 
```  
 

What we might guess from the above is that a graph of $(x,y_1)$ pairs
would show quite a lot of noise; $(x,y_2)$ pairs are a parabola with
some noise; $(x,y_3)$ pairs are for a straight line with some noise;
and, $(x,y_4)$ pairs are the same as the previous straight line, but
one point is some distance from the line.

The `WhereXY()` command for each plot is now given. 


```{r WhereXY} 
WhereXY(x,y2, grid=c(5,5)) 
WhereXY(x,y2, grid=c(5,5)) 
WhereXY(x,y3, grid=c(5,5)) 
WhereXY(x,y4, grid=c(5,5)) 
```  
 

We might also alter the number of cells in the grid to see how sensitive
we want to make our investigation. Personal preferences will determine
which combination of rows and columns works best. 

## Creation of graphs with an accessible interactive tool for exploration

New functions have been added to the `BrailleR` package in order to make
accessible tools for interactive exploration of graphs. The standard
commands for creation of histograms, boxplots, scatter plots and time
series plots do not return the extra information needed to get the
exploration tool to work. Each of the following commands creates the
same graph as would be created using the commands in previous chapters.
Refer to the examples found in the help pages for each of these commands
to see the standard base R commands that these `BrailleR` functions replicate.

N.B. You must store the graph in a named object for later use. 

```{r AccessGraphs1}
attach(airquality)
MyScatter = ScatterPlot(Wind, Ozone)
MyFLP = FittedLinePlot(Temp, Ozone, line.col=4)
MyHist = hist(Ozone)
MyBoxplot = boxplot(Wind)
detach(airquality)
``` 
 

## What else?

Future developments in the `BrailleR` package will be made to meet demand.
If you do want some additional functionality that you think is likely to
be wanted by others then please do add to the wish list. Current ideas
on the wish list include:  - Easier to interpret multiple regression output.  - Easier to interpret principal component analysis output.  - More graph types being described using the `VI()` command.  - Easier to interpret factor analysis output. 

```{r cleanup, include=FALSE, purl=FALSE}
rm(list = ls())
``` 
 

