The `detach()` command undoes the `attach()` command. To remove the “attachment", you will issue the command
`detach(airquality)`.

You could look at the help file for this data if you wanted to learn its
complete story using `?airquality`. It tells us that the data
are for daily readings of the following air quality values for May 1,
1973 (a Tuesday) to September 30, 1973 (a Sunday).  - `Ozone`: Mean ozone in parts per billion from 1300 to    1500 hours at Roosevelt Island  - `Solar.R`: Solar radiation in Langleys in the frequency    band 4000–7700 Angstroms from 0800 to 1200 hours at Central Park  - `Wind`: Average wind speed in miles per hour at 0700 and    1000 hours at LaGuardia Airport  - `Temp`: Maximum daily temperature in degrees Fahrenheit    at La Guardia Airport.

This data suits our purposes for the majority of examples in this
chapter but we will also need to look at another data set for the
discussion of graphs for discrete valued variables.

## Histograms

Like many graphical functions in R , the `hist()` command will
attempt to make a suitably attractive histogram with the minimum of
input from the user. Figure \@ref(fig:AirQuality-HistWind) shows what
results from the simplest application of the `hist()` command.

Note that the figure created has default settings for the main title,
axis labels and that the number of classes (also called bins) and the
cutoffs between them have been chosen automatically. Various methods
exist for these choices, but it is my recommendation that the user find
out what happens when the default settings are chosen and then alter
only what is actually necessary. For example, graphs in this document do
not always need the default title inserted so we need to suppress the
default action if we want to remove the title. We may also want more
informative axis labels. Both of these alterations are done for the
creation of Figure \@ref(fig:AirQuality-HistWind2).

```{r include=FALSE}
BlindOrSightedText = ifelse(Blind, "The need to know what is printed in a histogram was a primary motivator for the development of the BrailleR package and is one of the examples given in Section \\@ref(VIExplained) on providing text descriptions of graphs. The graphical version of this histogram is described in text using the tools of the `BrailleR` package; you can learn how this is actually done later.",
"")
```

`r BlindOrSightedText` 

```{r include=FALSE}
FigCap="A histogram showing the wind speed measurements."
```

