[Link to SVG version](../SVGs/TSPlotsCompared-1.html)

[Link to SVG version](../SVGs/TSPlotsCompared-2.html)
```{r include=FALSE}
BlindOrSightedText = ifelse(Blind, "### A note on the visual appearance of the standard time series plot (written for blind users)",
"")
```

`r BlindOrSightedText` `r if(Blind){Describe(TimeSeriesPlot(airquality$Wind))$general}` 

## Smoothing of a time series using moving averages {#MovingAverages}

One simple method for smoothing out a time series in order to expose its
behaviour is to smooth out the observed data using the
`filter()` command to create a moving average. A moving average
is often centred, (using the argument `sides=2` ) but if the
series is expected to have no trend, then a backwards only
(`sides=1` ) focus can be taken. We must determine the width
of the moving average, usually using odd-numbered widths.

Figure \@ref(fig:FilterExample) shows how different amounts of data can
be averaged to change the amount of smoothing.

```{r include=FALSE}
FigCap="Comparison of different amounts of smoothing a time
series."
```
